Staphylococcus aureus counters organic acid anion-mediated inhibition of peptidoglycan cross-linking through robust alanine racemase activity

Version 1

Abstract

Abstract

Weak organic acids are commonly found in host niches colonized by bacteria, and they can inhibit bacterial growth as the environment becomes acidic. This inhibition is often attributed to the toxicity resulting from the accumulation of high concentrations of organic anions in the cytosol, which disrupts cellular homeostasis. However, the precise cellular targets that organic anions poison and the mechanisms used to counter organic anion intoxication in bacteria have not been elucidated. Here, we utilize acetic acid, a weak organic acid abundantly found in the gut to investigate its impact on the growth of Staphylococcus aureus. We demonstrate that acetate anions bind to and inhibit D-alanyl-D-alanine ligase (Ddl) activity in S. aureus. Ddl inhibition reduces intracellular D-alanyl-D-alanine (D-Ala-D-Ala) levels, compromising staphylococcal peptidoglycan cross-linking and cell wall integrity. To overcome the effects of acetate-mediated Ddl inhibition, S. aureus maintains a high intracellular D-Ala pool through alanine racemase (Alr1) activity and additionally limits the flux of D-Ala to D-glutamate by controlling D-alanine aminotransferase (Dat) activity. Surprisingly, the modus operandi of acetate intoxication in S. aureus is common to multiple biologically relevant weak organic acids indicating that Ddl is a conserved target of small organic anions. These findings suggest that S. aureus may have evolved to maintain high intracellular D-Ala concentrations, partly to counter organic anion intoxication.

Organic acids produced by host and bacterial metabolism are critical determinants of infection outcomes (1, 2). During infection, the host macrophages produce millimolar amounts of itaconate, a dicarboxylic acid known to inhibit bacterial growth (3). Conversely, many bacterial pathogens and the gut microflora excrete short-chain organic fatty acids, which exhibit immunomodulatory functions and can skew the host response during infection (4, 5). Upon entry into the bacterial cell, organic acids can become toxic to bacteria when they disassociate in the cytosol as protons and organic anions. The influx of protons can result in cytoplasmic acidification and prove lethal for some pathogens if not adequately controlled (6). Similarly, organic anions have been shown to accumulate to toxic levels in the bacterial cytoplasm (7). However, the precise consequences of organic anion toxicity and the mechanisms pathogens employ to withstand the effects of anion perturbations within cells are not clearly understood.

Here we focus on the response of Staphylococcus aureus to acetic acid, which is the primary end-product of glucose catabolism under aerobic conditions. S. aureus also likely encounters high concentrations (up to 100 mM) of acetic acid and other short-chain fatty acids produced by human gut microbiota during intestinal colonization (8). On average, 20% of adults carry S. aureus in their intestines (9), and the burden there often surpasses that found in nasal passages by more than three orders of magnitude establishing the gut as a primary site for S. aureus colonization (10). We have previously shown that excreted acetic acid can promote cytoplasmic acidification in cultures of S. aureus, especially when the external environment becomes sufficiently acidic (pH< 5) (11). Cytoplasmic acidification promotes protein oxidation and triggers a staphylococcal ClpP-dependent damage response that eliminates unfit cells from the population (12). In contrast, in mildly acidic environments (pH 5.5-6.5), although S. aureus actively buffers its intracellular environment against acidification, the transmembrane pH gradient (ΔpH) of S. aureus will drive the accumulation of millimolar quantities of acetate anions into the cytoplasm. Previous studies in Escherichia coli have shown that acetate intoxication causes an osmotic imbalance that can transiently be accommodated by the efflux of physiological anions like glutamate (7). In addition, acetate anions have also been reported to impact enzymes in the methionine biosynthetic pathway, resulting in a toxic accumulation of homocysteine and a reduction in intracellular methionine leading to growth inhibition of E. coli (13, 14). However, it remains unclear if these effects are common to other bacteria.

Here, we demonstrate that the primary target of acetate intoxication in S. aureus is Ddl. This crucial enzyme produces the D-Ala-D-Ala dipeptide, that is incorporated into peptidoglycan precursors and is necessary for cell wall cross-linking. We also demonstrate that carbon flux through alanine racemase and a tight control of Dat activity increases the cytosolic D-Ala pools to counter acetate-mediated inhibition of Ddl. Importantly, these phenotypic effects are not unique to acetate but are conserved across multiple biologically important organic acid anions. Therefore, we propose that S. aureus may have evolved to maintain a high intracellular D-ala pool partly to offset the inhibition of Ddl by organic anions typically encountered during human colonization.

Results

Alanine racemase counters acetate intoxication

To identify genetic determinants that counter the effects of acetate intoxication, we screened the Nebraska Transposon Mutant Library (NTML) for mutants sensitive to 20 mM acetic acid in Tryptic Soy Broth (TSB) media, pH 6.0. Under these conditions, S. aureus maintains its intracellular pH approx. 1.5 units above the external pH (15) and is estimated to accumulate over 600 mM acetate in the cytosol (16). The NTML strains were grown under static conditions at 37°C, and the extent of growth was determined at 24 h by measuring the optical density at 600 nm (OD600). As a control, we performed an identical screen without acetic acid supplementation. We normalized the growth of each mutant in both screens (± acetic acid) to their isogenic wild-type (WT) strain. A comparison of growth indices (OD600 Tn-mut/WT) for each mutant in the presence and absence of 20 mM acetic acid revealed that most mutants clustered close to an index of 1 in the plot (Figure 1A), which suggested that most mutants tolerated acetate intoxication reasonably well. A few mutants that grew poorly following acetate intoxication due to inherent growth defects were observed close to the plot diagonal, whereas those mutants that did not have intrinsic growth deficiencies were located further away from the diagonal. Among the latter class of mutants, we observed that the alr1 mutant had the most substantial reduction in growth when subjected to acetate stress (Figure 1A, B). To confirm that the acetate-dependent growth defect of the alr1 mutant was not due to polar effects, we complemented the mutant by inserting a functional copy of alr1 under the control of its native promoter into the S. aureus pathogenicity island (SaPI) attachment site. Genetic complementation completely restored the alr1 mutant phenotype to WT levels (Figure 1B). These results suggest that acetate intoxication impairs the growth of S. aureus in the absence of a functional alanine racemase. Further supporting this conclusion, we could reduce acetate toxicity in the alr1 mutant by culturing this strain in glucose-free TSB media, which alleviates carbon catabolite repression and activates TCA cycle-dependent acetate catabolism (Figure 1C) (17). Conversely, the inactivation of citrate synthase (citZ), the first enzyme of the TCA-cycle, re-imposed acetate toxicity in the alr1 mutant when cultured in glucose-free TSB media (Figure 1C).

Alanine racemase activity counters acetate intoxication

(A) The Nebraska Transposon Mutant library was screen against 20 mM acetic acid, pH 6.0 to identify mutants with altered growth phenotypes. The WT strain and transposon mutants were grown for 24 h in TSB ± 20 mM acetic acid. The bacterial growth at 24 h was measured spectrophotometrically (OD600) and normalized to WT growth. The X and Y-axis on the plot represent normalized growth values for each mutant in the presence or absence of acetate. (B) The growth of the WT, alr1 mutant, and alr1 complemented strain in TSB supplemented with 20 mM acetic acid. (C) Aerobic growth of WT, alr1, citZ, citZalr1 mutants in TSB media lacking glucose but supplemented with 20 mM acetic acid. (D) LC-MS/MS analysis was performed to quantify the intracellular D-Ala-D-Ala pool in strains cultured for 3 h (exponential phase) in TSB ± 20 mM acetic acid. (E) The growth of strains was monitored following D-Ala supplementation (5 mM) in TSB + 20 mM acetate, (n=3, mean ± SD). Ac, acetate.

Acetate intoxication alters the intracellular D-Ala-D-Ala pools

Alr1 catalyzes the conversion of L-Ala to D-Ala during staphylococcal growth (Figure 1-figure supplement 1A). The D-Ala is further converted to D-Ala-D-Ala dipeptide by the ATP-dependent Ddl (Figure 1-figure supplement 1A) and incorporated into peptidoglycan (PG) muropeptide, thus playing a crucial role in PG biosynthesis, cross-linking, and integrity (18, 19). Therefore, we hypothesized that under acetate stress, low concentrations of D-Ala in the alr1 mutant might concomitantly reduce D-Ala-D-Ala concentrations in the cell resulting in a growth defect. To test this hypothesis, we determined the intracellular pool of D-Ala-D-Ala using liquid chromatography-tandem mass spectrometry (LC-MS/MS). In regular growth media (TSB), we observed that the inactivation of alr1 decreased the D-Ala-D-Ala pool by approximately 80% compared to the WT strain (Figure 1D). However, following acetate intoxication the level of D-Ala-D-Ala was depleted by more than 99% (Figure 1D). The external supplementation of D-Ala (5 mM) in the media fully restored the growth of the alr1 mutant to WT levels under acetic acid stress (Figure 1E), which suggests that increased intracellular D-Ala pools can overcome the detrimental impact of acetate intoxication.

The depletion of D-Ala-D-Ala following acetate intoxication is surprising since S. aureus is predicted to have two additional pathways that can synthesize D-Ala and channel it to the production of this dipeptide. For instance, S. aureus harbors a second predicted alanine racemase (Alr2) that could compensate for the lack of Alr1 activity (Figure 1-figure supplement 1A). Alternatively, Dat, which catalyzes the formation of D-Ala from pyruvate and D-glutamate (D-Glu), may functionally complement the alr1 mutant under acetate stress (Figure 1-figure supplement 1A). However, the lack of functional complementation from these alternate pathways of D-Ala biosynthesis following acetate intoxication suggests that not all metabolic routes to D-Ala are operational or that regulatory bottlenecks limit pathway activity. To test these possibilities, we constructed a series of mutants in which all three predicted routes of D-Ala biosynthesis (alr1, alr2 and dat) were disrupted either individually or in various combinations and performed growth assays (Figure 1-figure supplement 1B). Surprisingly, we observed that the inactivation of alr1 and dat simultaneously (alr1dat mutant) was synthetic lethal in S. aureus, suggesting that alr1 and dat were the sole contributors of D-Ala in S. aureus. Indeed, the supplementation of D-Ala fully restored the growth of the alr1dat mutant (Figure 1-figure supplement 1C).

The inactivation of alr2, either alone or in combination with other D-alanine-generating enzymes, did not affect growth (Figure 1-figure supplement 1B). This observation suggests that alr2 is unlikely to be a functional alanine racemase under the growth conditions tested. Collectively, these results indicate that Dat activity accounts for D-Ala production in the absence of alr1, but its contribution is insufficient to counter acetate intoxication.

Insufficient translation of dat impacts the alr1 mutant following acetate intoxication

Since Dat activity contributes to D-Ala production in the alr1 mutant, we questioned why Dat is insufficient to sustain D-Ala-D-Ala pools under conditions of acetate intoxication. One possible explanation may relate to the maintenance of osmotic balance by S. aureus. It has been proposed that the intracellular accumulation of acetate anions may bring about an efflux of L/D-Glu from cells to adjust for osmolarity, thus exhausting one of the key substrates for Dat activity and limiting D-Ala production (7). However, this hypothesis is improbable since the expression of dat from a multicopy vector rescued the alr1 mutant from the effects of acetate intoxication (Figure 2-figure supplement 1A), suggesting that the intracellular D-Glu pools are sufficient to support D-Ala production through Dat activity. Alternatively, we hypothesized that the alr1 mutant’s heightened sensitivity to acetate toxicity could be due to a decrease in dat transcription which would effectively reduce intracellular D-Ala. However, we found no detrimental effect of acetate intoxication on dat transcription in the alr1 mutant (Figure 2-figure supplement 1B). Together, these observations raise the possibility that the depletion of D-Ala-D-Ala in the alr1 mutant following acetate intoxication may arise from a post-transcriptional regulatory bottleneck that limits dat from meeting the demand for intracellular D-Ala.

In S. aureus, dat is part of a bicistronic operon (Figure 2A). The first gene, pepV, encodes an extracellular dipeptidase (20, 21). Transcriptional start site (TSS) mapping of the pepV-dat operon by the adaptor and radioactivity-free (ARF-TSS) method revealed a 30-nucleotide untranslated region (5’-UTR) extending upstream from the pepV initiation codon. The 5’-UTR includes a Shine-Dalgarno motif (ribosome binding site, SD1) upstream of the pepV start codon (Figure 2A). In addition, a second SD motif (SD2) associated with dat was identified within the pepV coding region (Figure 2A) and did not overlap with the pepV termination codon. The location of SD2 within pepV suggests that the insufficient production of D-Ala by dat following acetate intoxication could be attributed to suboptimal translation of dat. This could occur as ribosomes (70S) that are moving from SD1 may interfere with the translation of dat from SD2. To test this hypothesis, we engineered a nonsense mutation in pepV (alr1pepVQ12STOP mutant) that would prevent the ribosomes originating from SD1 from moving forward (Figure 2A). However, the alr1pepVQ12STOPmutant grew poorly compared to the alr1 mutant following acetate intoxication (Figure 2B). This suggested that the translation of dat is coupled to that of pepV presumably through stable mRNA secondary structures that form within pepV. These structures may not be effectively resolved in the alr1pepVQ12STOPmutant due to the absence of ribosome traffic on pepV mRNA.

Translational coupling of dat to pepV limits the alr1 mutant from countering acetate intoxication

(A) Schematic representation of various engineered mutations in the pepV-dat locus.SD, Shine-Dalgarno motif; TSS, transcriptional start site. (B)-(D) Growth of engineered mutants was monitored spectrophotometrically (OD600) in TSB supplemented with 20 mM acetate (n=3, mean ± SD). (E) RT-qPCR to determine dat transcription in various mutants relative to the WT strain.

As an alternative approach to determine if SD2 positioning within pepV impeded dat translation, we deleted pepV along with SD1 in the alr1 mutant (alr1pepVΔSD1-467, Figure 2A). In the resulting strain, dat translation was under the sole control of its native SD2. Remarkably, the alr1pepVΔSD1-467 mutant did not display a heightened sensitivity to acetate stress and grew identical to the WT strain following acetate intoxication (Figure 2C). Similarly, an alr1 mutant in which dat was linked to SD1 (alr1pepV mutant, Figure 2A) also phenocopied the WT strain following acetate intoxication (Figure 2D). Notably, the observed growth differences in alr1pepVΔSD1-467, alr1pepV and alr1pepVQ12STOP mutants following acetate intoxication did not result from any changes in dat transcription (Figure 2E). These findings collectively suggest that the native promoter elements, as well as the SD sites of pepV and dat can independently support the robust expression and translation of dat to levels required for countering acetate intoxication. However, the genetic arrangement of the dat translation initiation region (TIR) within pepV offered tight control of dat translation and prevented cells from producing sufficient enzyme following acetate intoxication.

Why is the Dat tightly controlled?

The need to tightly control Dat activity suggests that flux between D-Ala and D-Glu pools must be carefully balanced during staphylococcal growth. To gain insight into this process, we profiled the mass isotopologue distribution (MID) of D-Ala-D-Ala in the WT, alr1, and dat mutants under isotopic steady-state conditions using 13C 15N-L-Ala as the tracer during growth experiments in chemically defined medium (CDM). The flux of 13C 15N-L-Ala through Alr1 should result in 13C 15N-D-Ala production (Figure 3A, D-Ala retains labeled nitrogen). On the other hand, staphylococcal alanine dehydrogenases (Ald1 and Ald2) catalyze the conversion of 13C 15N-L-Ala to 13C-pyruvate and finally 13C-D-Ala through Dat activity (Figure 3A). Thus, the labeled nitrogen in 13C 15N-L-Ala is lost as 15N-NH when fluxed through the Ald/Dat pathway (Figure 3A). Since the intracellular pools of D-Ala are converted to D-Ala-D-Ala, the MID of the latter metabolite should mirror the isotopologue ratios of D-Ala produced from either Alr1 or Dat activities.

Reaction orientation and fluxes through Alr1 and Dat

(A) Schematic representation of various isotopologues of D-Ala-D-Ala and D-Glu generated from 13C 15N 3 1 labeled L-Ala. Metabolites in blue mainly arise from Alr1, red, through the Ald1/2-Dat pathway and yellow are unlabeled intermediates within cells. The mass isotopologue distribution of (B) D-Ala-D-Ala and (C) D-Glu were determined by LC-MS/MS following the growth of S. aureus in chemically defined media supplemented with 13C 15N1 L-Ala (n=3, mean ± SD). Isotopologues of D-Ala-D-Ala shown in grey color are minor species and are noted in Table S9. (D) The mean competitive fitness (w) was determined by co-culturing the WT strain with an isogenic mutant that contained either the empty pAQ59 vector integrated into the SaPI chromosomal site (WTEV) or the pAS8 vector containing dat under the control of its native promoter (WTdat) (n=18, the dotted lines indicate the median and quartiles).

LC-MS/MS analysis revealed that ∼80% of the intracellular D-Ala-D-Ala pool had incorporated the labeled L-Ala supplemented in media (fractional contribution, 0.80). As expected, the majority (∼55%) of the D-Ala-D-Ala in the WT was composed of the C6N2 isotopologue (in which both units of D-Ala contain labeled carbon and nitrogen), which suggested that alr1 was the major contributor of D-Ala in S. aureus (Figure 3B). Surprisingly, the sole contribution of dat activity (C6N0, C3N0, C0N0) to D-Ala-D-Ala was less than 1% in the WT strain and D-Ala-D-Ala isotopologues with at least one D-Ala originating from dat activity (C6N1, C3N1, C3N2) although readily observed, were still in the minority. However, the D-Ala-D-Ala originating from Dat activity expanded substantially upon alr1 mutation (Figure 3B). Inactivation of dat itself displayed few differences in the MID of D-Ala-D-Ala, compared to the WT strain (Figure 3B). These results suggest that flux through Dat is most likely driven towards D-Glu in the WT strain rather than D-Ala. Only upon inactivation of alr1 does the Dat activity reverse towards the production of D-Ala.

To confirm these predictions, we measured the levels of 15N1-D-Glu in the WT, alr1, and dat mutants following growth with the 13C 15N-L-Ala tracer. Consistent with Dat activity funneling D-Ala to D-Glu in the WT, approximately 78% of the D-Glu pool in the WT strain was 15N labeled. Furthermore, we observed that inactivation of dat resulted in the complete depletion of intracellular levels of 15N1-D-Glu (Figure 3C). Inactivation of alr1 also had a similar outcome with loss of 15N1-D-Glu pools due to the lack of 13C 15N-D-Ala in this mutant (Figure 3C). Together, these results strongly suggest that in the WT strain, Dat activity diverts D-Ala towards D-Glu production.

Given the critical need to produce D-Ala-D-Ala during acetate intoxication, any diversion of its precursor pool (D-Ala) to D-Glu through Dat activity is bound to decrease cell fitness and thus may justify its tight translational control. To test this hypothesis, we determined the mean competitive fitness (w) of cells that overexpressed dat compared to those that had native levels of expression. Accordingly, we performed coculture competition assays of the WT strain with an isogenic mutant strain that either harbored an empty vector (pAQ59) integrated into the SaPI chromosomal site or a vector containing dat under control of its native promoter (pAS8), following acetate intoxication. Consistent with increased Dat activity in the WT strain being detrimental to the cell, the mean competitive fitness of the dat overexpressing strain was significantly lower (w4h= 0.91) in the exponential growth phase than its isogenic WT strain that harbored the empty vector (w4h= 1.26) (Figure 3D). Collectively, these results suggest that Dat catalyzes the production of D-Glu in the WT strain, and its tight regulation prevents excessive flux of D-Ala to D-Glu which is necessary to maintain cell fitness following acetate intoxication.

Acetate intoxication impacts PG biosynthesis

Since acetate intoxication ultimately affects D-Ala-D-Ala pools (Figure 1D), we predicted potential alterations to PG biosynthesis and cell wall integrity. To test this hypothesis, we quantified various cytosolic PG intermediates in the WT strain by LC-MS/MS analysis. Acetate intoxication caused a significant increase in the intracellular pools of multiple PG biosynthetic intermediates, including Uridine diphosphate N-acetylglucosamine (UDP-NAG), UDP-N-acetylmuramic acid (UDP-NAM), UDP-NAM-L-Ala, UDP-NAM-L-Ala-D-Glu-L-Lys and UDP-NAM-L-Ala-D-Glu-L-Lys-D-Ala-D-Ala (UDP-NAM-AEKAA) in the WT strain when compared to the unchallenged control (Figure 4A). However, the growth of the WT strain was slightly inhibited by acetic acid (Figure 1B), which suggests that the observed accumulation of PG intermediates may have been due to an imbalance between the rates of PG biosynthesis and growth. Notably, the alr1 mutant showed higher levels of UDP-NAM-AEK compared to the WT and the dat mutant following acetate intoxication (Figure 4A), indicating a metabolic block in the production of UDP-NAM-AEKAA due to insufficient D-Ala-D-Ala. The effect of this metabolic block is also evident from the increased transcription of ddl and murF (Figure 4B) which encode enzymes that incorporate D-Ala-D-Ala into PG precursors, suggesting a greater need to maintain peptidoglycan cross-linking following acetate intoxication.

Acetate intoxication impacts soluble PG precursor pools and cell wall cross-linking.

(A) The intracellular pool of PG intermediates in exponential phase cultures of S. aureus was estimated using LC-MS/MS analysis. cps, counts per second (B) ddl and murF transcription in the exponential growth phase was determined by RT-qPCR analysis (n=3, mean ± SD). (C) Cell wall muropeptide analysis of the WT, alr1 and dat mutants was determined following growth in TSB ± 20 mM acetate for 3 h. Cell wall cross-linking was estimated as previously described (49). Ac, acetate.

Unsurprisingly, the dysregulation of D-Ala-D-Ala homeostasis following acetate intoxication was also reflected in the extent of cell wall cross-linking in the WT, alr1 and dat mutants. Muropeptide analysis revealed that acetate intoxication in the WT strain increased levels of monomeric muropeptides (Figure 4C, Figure 4-figure supplement 1). Conversely, the percentage of di- and trimeric muropeptides decreased relative to the WT control, as did the percent cross-linking (Figure 4C). These observations suggest that acetate intoxication constrains the D-Ala-D-Ala pool in the WT strain and alters PG cross-linking despite Alr1 activity. The extent of PG cross-linking in the dat mutant was similar to WT in the presence or absence of acetate, consistent with our finding that the Dat activity plays a limited role in maintaining the D-Ala-D-Ala pool in the WT strain (Figure 4C). In contrast, PG cross-linking in the alr1 mutant was lower than the WT strain by ∼10% (Figure 4C). Acetate intoxication further decreased the cross-linking approximately 20% relative to WT as well as the ratio of dimeric to monomeric muropeptides in the alr1 mutant, which inevitably reduced the growth of this strain (Figure 4C).

Muropeptide analysis also revealed the accumulation of a disaccharide tripeptide (NAG-NAM-AEK (M3); m/z. Da, 826.4080) in the peptidoglycan (PG) extracted from the alr1 mutant (Figure 4-figure supplement 1B). This finding suggests that the significantly elevated levels of UDP-NAM-AEK in the alr1 mutant could efficiently outcompete the substrate specificity of phospho-N-acetylmuramyl pentapeptide translocase (MraY) for UDP-NAM-AEKAA, ultimately becoming integrated into the PG structure itself. Interestingly the incorporation of UDP-NAG-NAM-AEK into the alr1 mutant’s PG only marginally increased following acetate treatment (Figure 4-figure supplement 1B, see inset). The increase of UDP-NAG-NAM-AEK is most likely an underestimate since cells with higher levels of incorporation are more likely to lyse due to a reduction in PG cross-linking. Overall, these observations support a model wherein the immediate consequences of acetate intoxication are defects in PG crosslinking and biosynthesis.

Acetate intoxication inhibits Ddl activity

While the above observations point to the consequences of acetate intoxication of S. aureus, its molecular target was not initially identified. Since acetate intoxication dramatically reduces D-Ala-D-Ala levels in the alr1 mutant (Figure 1D), we reasoned that acetate might inhibit either Dat or Ddl activity. To distinguish between these two targets, we measured the levels of D-Ala in the alr1 mutant following acetate intoxication. Surprisingly, we observed that the D-Ala pools in the alr1 mutant did not significantly change in response to acetate intoxication compared to the untreated control (Figure 5A). This suggested that Dat activity was preserved in the alr1 mutant to the same extent as its untreated control and was not affected by acetate.

Acetate anion inhibits Ddl activity.

(A) The intracellular D-Ala was determined by LC-MS/MS analysis. (B) The ddl gene was overexpressed in S. aureus using a cadmium inducible expression system (pSP36). CdCl2, 0.312 µM. (C) Inhibition of recombinant His-tagged Ddl activity in the presence of 300 mM sodium acetate (D) IC50 curve of the inhibition of rDdl by acetate. Michaelis-Menten kinetics of rDdl in varying concentrations of (E) D-Ala, and (F) ATP in the presence of acetate to assess the inhibition mechanism. (G) Structure of the acetate bound Ddl (PDB:8FFF). (H) Acetate bound to the ATP binding site of Ddl (I) Acetate bound to the second D-Ala binding site of Ddl. The calculated Fo-Fc omit maps are contoured to 3σ and the mesh is shown in blue. (J) Superimposed structure of acetate bound Ddl (slate blue) with StaDdl apo structure (PDB:2I87, beige) and StaDdl-ADP complex structure (PDB:2I8C, grey) showing a shift of ω loop (red) to ATP binding site. The D-Ala-D-Ala was modeled at the D-Ala binding site using Thermos thermophius HB8 Ddl structure (PDB:2ZDQ). The bound ADP (grey) of PDB:2I87 and modeled D-Ala-D-Ala (light blue) indicates the positioning of Ac at ATP and second D-Ala binding sites respectively. Ac, acetate; V, velocity.

Conversely, these findings also indicate that the acetate-dependent decrease of the D-Ala-D-Ala pool in the alr1 mutant was most likely due to the inhibition of Ddl. To test this hypothesis, we cloned S. aureus ddl under the control of a cadmium inducible promoter and induced its expression in the alr1 mutant following acetate intoxication (Figure 5B). Indeed, the growth of the alr1 mutant was restored to WT levels when ddl was overexpressed, strongly suggesting that Ddl was the target of acetate anion (Figure 5B).

To confirm that acetate inhibits Ddl through direct interactions, we undertook two separate approaches. As the first approach, 6xHis-tagged S. aureus Ddl was purified, and in-vitro enzyme kinetic assays were performed to determine the possible inhibitory mechanism of Ddl by acetate. Considering the high concentration of acetate estimated to accumulate in the cytoplasm, a concentration of 300 mM sodium acetate was used in the initial reactions to test inhibition (Figure 5C). Interestingly, variation of acetate concentration showed that Ddl was inhibited in vitro, and these conditions suggest an IC50 of 400.3 ± 8 mM (Figure 5D). This indicates significant inhibition of Ddl by acetate when the cellular concentration is near the hypothesized 600 mM, further confirming that Ddl is a direct target of inhibition by acetate anion. Furthermore, based on kinetic experiments performed under varying concentrations of either ATP or D-Ala, the kcat values are shown to be distinctly different for each acetate concentration, which strongly suggests a mixed inhibition mechanism for acetate (Figure 5E and F, Table S1).

Differential Scanning Fluorometry (DSF) was used as another approach to assess the direct binding of acetate to Ddl (Table S2). The Ddl protein without any ligand bound shows a melting temperature (Tm) of 45 °C. After adding 300 mM sodium acetate, Ddl exhibited a 3.7 °C Tm shift indicating a slight thermal stabilization upon binding acetate. This is higher than the shift in the Tm exhibited by a Ddl/ATP complex. The addition of ADP to Ddl results in a decrease of 3.2 °C, indicating a decrease in thermal stability compared to Ddl alone. Intriguingly, when adding acetate to Ddl complexes with ATP or ADP, the Tm increased to 48.9 °C and 49.9 °C, respectively (Table S2). This represents a Tm increase of 2.3 °C when acetate is added to a Ddl/ATP complex but a Tm increase of 8.1 °C when acetate is added to a Ddl/ADP complex. The addition of D-Ala to the reaction mixture increases the Tm of Ddl by 4.2 °C and adding acetate to the Ddl/D-Ala mixture shows only a 0.3 °C Tm shift (Table S2). The widely varying changes in Tm for the tested complexes, particularly when comparing the Tm values for ligand-free, ADP-bound, and the ADP/Acetate complex, further support a mixed inhibition mechanism as these data suggest acetate may bind to multiple sites on Ddl or the location of these binding sites may change depending on the ligand-bound state of the enzyme due to Ddl conformational changes as observed in Ddl orthologs (22).

Ddl/Acetate complex structure shows binding of acetate at both substrate binding sites

To gain further insight into the mechanism of acetate inhibition, the X-ray crystal structure of a Ddl/acetate complex was obtained using co-crystals of Ddl and acetate. The crystal diffracted to 1.9 Å and data were consistent with a P 2 21 21 space group possessing one molecule of Ddl in the asymmetric unit (Table S3). The crystal structure of the Ddl/acetate complex (PDB:8FFF) shows difference density corresponding to acetate at two different sites of the protein. One acetate is positioned within the adenine binding subsite of the ATP binding site and the other acetate ion is positioned in the second D-Ala binding site (Figure 5G). The acetate ion in the ATP binding site interacts with the side chain of Lys177 and the backbone nitrogen of Val216. Also, the methyl group of acetate forms van der Waals interactions with the side chain of Leu145 (Figure 5H). The acetate ion that binds to the D-Ala binding site forms a bidentate polar interaction with the side chain of Arg291 and a hydrogen-bonded interaction with the backbone nitrogen of Gly312 (Figure 5I). These two residues are conserved in Ddl homologs and previous structural data clearly illustrate the crucial role these residues play in D-Ala binding (22).

The acetate-bound structure shows conformational differences compared to the previously published ligand-free and ADP-bound structures (22). The ω loop, which is associated with substrate binding, is disordered in both the S. aureus Ddl ligand-free (PDB:2I87) and the Ddl-ADP complex structures (PDB:2I8C) as well as other available Ddl crystal structures that lack bound substrates or ligands (PDB:3K3P, 5DMX and 6U1C) (23–25). Interestingly, this loop is well ordered in the acetate-bound structure described here (Figure 5J), which gives the first view of the S. aureus Ddl ω loop and the interactions it may form with substrates or inhibitors. The structural stabilization of the ω loop is consistent with the DSF results exhibiting an increase in the melting temperature upon binding acetate. The ω loop is shifted towards the ATP binding site and repositions the conserved Tyr246 side chain within the ATP binding site, which likely hinders the binding of ATP (Figure 5J). This positioning is comparable with the Mycobacterium tuberculosis Ddl (PDB:3LWB) ligand-free structure, which also takes a closed conformation showing the ω loop positioned within the ATP binding site and obstructing ATP binding (26). Taken together, the kinetic, DSF, and structural data suggest that while acetate can directly bind within both substrate binding pockets of Ddl, it also stimulates conformational changes in the dynamic ω loop to afford more allosteric-like effects on enzyme activity. Each of these observations support a mixed inhibition modality.

Multiple organic acids inhibit the alr1 mutant in a D-Ala-dependent manner

Finally, we determined whether the growth inhibition of the alr1 mutant is unique to acetate anion or is a more general phenomenon mirrored by addition of other small organic acids. Accordingly, we initially performed molecular docking studies of three biologically relevant organic anions: lactate, propionate and itaconate, in both the ATP and D-Ala binding pockets of Ddl (Figure 6A-D). The acetate anion-bound structure of Ddl was used as a reference for analysis. The docking results suggest reasonable poses for lactate, propionate and itaconate within the ATP binding site forming polar interactions with Ddl residues conserved for binding ATP. Upon docking, the carboxylate moieties of both lactate and propionate form ionic interactions with the Lys177 side chain similar to those observed in the Ddl/acetate crystal structure (Figure 6A and B). Also, the side chain of Glu213 in Ddl forms a hydrogen bonded interaction with the hydroxyl of lactate (Figure 6A) and van der Waals interactions between propionate and nearby side chains of Phe175 and Phe295 (Figure 6B) were indicated. The two carboxylate groups of itaconate form hydrogen bonded interactions with backbone amide nitrogen atoms of Ala218 and Tyr246 as well as a van der Waals interaction with the nearby side chain of Phe175 (Figure 6C).

Biologically relevant weak acids inhibit growth of the alr1 mutant.

Molecular docking of (A) lactate (B) propionate and (C) itaconate to the ATP binding site of Ddl. (D) The relative positions and poise of different organic anions in relation to acetate in the D-Ala binding site of Ddl was determined using Schrödinger Glide. The growth (OD600) of the WT and alr1 mutant in TSB containing (E) lactic acid (40 mM) (F) propionic acid (20 mM) and (G) itaconic acid (20 mM) in the presence or absence of 5 mM D-Ala.

The molecular docking results for lactate, propionate, and itaconate in the D-Ala binding site of Ddl also show similar types of interactions but with variable poses and slight orientation differences compared to that observed for acetate in the crystal structure (Figure 6D). The D-Ala binding site, consisting of primarily charged and polar atoms, allows for a range of binding modes for these small anions, where the ligand size is a stronger factor in determining the binding location. Acetate and propionate, being smaller and less sterically hindered, bind preferentially near the Arg 291 side chain that coordinates the acid moiety of D-ala during the enzymatic reaction (Figure 6D). Meanwhile, itaconate and L-lactate bind in the more spacious region between Lys251 and Ser317 (Figure 6D). The Glide scores from the docking results, which provide a rough estimate of the ΔG of binding for each ligand suggest modest affinity to the identified binding sites (Table S4).

To determine if these organic acids could impact Ddl function, the WT and the alr1 mutant were challenged with lactic, propionic and itaconic acids (Figure 6E-G). All three organic acids inhibited the growth of the alr1 mutant. The addition of D-Ala to the culture media rescued the growth of the alr1 mutant to WT levels (Figure 6E-G) consistent with Ddl being the target of lactate, propionate and itaconate. Moreover, overexpression of ddl in the alr1 mutant also restored growth of the alr1 mutant following the organic acid challenge (Figure 6-figure supplement 1A-C). These findings collectively suggest that various organic acid anions can inhibit Ddl activity in S. aureus.

Discussion

Intracellular anion accumulation has long been hypothesized to drive weak organic acid toxicity in bacteria (16, 27, 28). However, few studies have investigated the mechanism by which weak acid anions inhibit bacterial growth. Acetic acid is particularly interesting among weak acids, given that it is a common byproduct of glucose catabolism in bacteria and is excreted in high concentrations (29). S. aureus does not catabolize acetate as a carbon source unless glucose is first exhausted from the environment (30). This results in the intracellular accumulation of acetate in S. aureus as a function of the bacterial transmembrane pH gradient, especially when acetic acid concentrations are high in the immediate vicinity of cells. Here we determine that at high intracellular concentrations, acetate anions directly bind Ddl and inhibit D-Ala-D-Ala production to adversely impact peptidoglycan cross-linking (Figure 7). However, S. aureus exhibits a remarkable tolerance to acetate intoxication due to the robust production of D-Ala by Alr1, which ultimately increases D-Ala-D-Ala pools (Figure 7).

Model depicting the role of Alr1 in countering organic acid anion-mediated inhibition of Ddl.

During its growth, S. aureus (WT) maintains a substantial intracellular pool of D-Ala through the activity of Alr1. Any excess D-Ala is subsequently converted into D-Glu by the action of the Dat enzyme. The high concentration of D-Ala is crucial for the optimal functioning of Ddl and serves to prevent the inhibition of Ddl by acetate (Ac-) and other organic acid anions. This process generates sufficient D-Ala-D-Ala, which is rapidly incorporated into the PG tripeptide precursor UDP-NAM-AEKAA to form UDP-NAM-AEKAA, which ultimately contributes to a robust cross-linked PG (murein) sacculus. In the alr1 mutant, the Dat reaction orientation is switched to preserve intracellular D-Ala. Nevertheless, this change is inadequate to maintain sufficient D-Ala pool to shield Ddl from inhibition by Ac-, due to tight control of dat translation. This results in an excess of UDP-NAM-AEK, which competes effectively with UDP-NAM-AEKAA for PG incorporation. The absence of a terminal D-Ala-D-Ala in the PG hinders crosslinking and leads to impaired growth following acetate intoxication.

Multiple lines of evidence demonstrate Ddl to be the target of acetate anions. First, LC-MS/MS analysis revealed that acetate intoxication decreased D-Ala-D-Ala pools but not D-Ala in S. aureus, pointing to Ddl as the target of acetate. Second, DSF and in-vitro enzyme kinetic studies showed that acetate could bind and inhibit purified rDdl through a mixed inhibition mechanism. Third, structural analysis of the Ddl-inhibitor complex confirmed that acetate binds to both the ATP-binding and D-Ala binding sites within Ddl and further induced conformational changes to the dynamic ω loop, which weakens the binding of ATP to the Ddl active site. Finally, overexpression of ddl alone was sufficient to overcome acetate-mediated inhibition of the alr1 mutant and restore growth to WT levels.

Inhibitors that bind an enzyme’s catalytic substrate binding sites are usually competed out by high concentrations of substrates. However, acetate inhibits Ddl through a mixed inhibition mechanism despite binding to the substrate binding pockets of Ddl. We suspect this is due to additional conformational changes observed in the dynamic ω loop that affords more allosteric-like effects on enzyme activity. However, we cannot rule out that acetate might bind to additional sites in the Ddl-ATP complex, Ddl-ADP complex, or a Ddl-ADP-phospho-D-Ala complex with varying affinities. The differences in the temperature shifts observed in DSF with various substrate complexes support this possibility. The crystal structures of Ddl/acetate complexes with different substrates could provide a more precise conclusion about the inhibitory modality of Ddl by acetate. In line with acetate’s inhibitory effect on Ddl, we observed that acetate intoxication in the alr1 mutant led to a disproportionate increase in the cytosolic pool of PG tripeptide intermediate (UDP-NAM-AEK) compared to the pentapeptide form (UDP-NAM-AEKAA). Previous reports have suggested that MraY might facilitate the integration of UDP-NAM-tripeptide into S. aureus PG, especially when its concentration within cells exceeds that of UDP-NAM-pentapeptide (31, 32). Our findings strongly support this hypothesis, as the analysis of the alr1 mutant’s cell wall muropeptides revealed a clear elevation in the level of the disaccharide-tripeptide NAG-NAM-AEK. The inhibition of Ddl by acetate would further reduce the presence of terminal D-Ala-D-Ala moieties within alr1 muropeptides which likely leaves these cells incapable of withstanding the outward-directed cell turgor pressure, ultimately leading to cell death (32).

Despite acetate inhibiting Ddl through a mixed inhibition mechanism, it should be noted that a functional Alr1 or even the supplementation of D-Ala in culture media can provide significant tolerance against acetate intoxication in S. aureus. These observations suggest that Ddl is only weakly inhibited by acetate, which is also evident from the relatively high IC50 of approximately 400 mM observed in our kinetic experiments with S. aureus Ddl. The weak inhibition of Ddl would suggest that inflating the cytosolic D-Ala pools could promote sufficient generation of D-Ala-D-Ala to counter acetate intoxication. Indeed, it has been estimated that S. aureus maintains a high concentration of roughly 30 mM intracellular D-Ala (33), which we now demonstrate to be critical in countering acetate intoxication.

The existence of pepV and dat within the same operon suggests that these genes may have evolved related functions. In Lactococcus lactis the PepV dipeptidase activity was shown to be important for supplying cells with L-Ala which was eventually incorporated into PG (34). In this context, pepV and dat may have a similar role in modulating the intracellular alanine pool. A surprising finding of our study was that dat expression is relatively stable and tightly controlled in S. aureus due to its SD motif being located within the coding region of pepV.

Furthermore, such a genetic arrangement has been linked to translational coupling (35), wherein active translation from the first gene promotes the translation of the following gene in the operon, which in the case of dat was not sufficient to overcome acetate toxicity in the alr1 mutant. Two central mechanisms of translational coupling have been proposed. The first involves secondary and tertiary mRNA structures that either occlude or encompass the SD motif of downstream genes and shield it from ribosomes, thus preventing its translation (36). These mRNA structures can be relieved when a ribosome initiates translation from the first gene of the operon and exposes the downstream intragenic SD sequences to new 30S ribosomal subunits (34). In the second mechanism, continued translation of the first gene of the operon is necessary to increase the abundance of ribosomes in the TIR of the second gene resulting in its enhanced translation (35). Irrespective of the mechanism of translational coupling, our results suggest that genetic arrangements that promote translational coupling might also limit the overall production of dat and thus prevent it from functionally complementing the alr1 mutant following acetate intoxication. Since our data suggest that Dat primarily promotes flux from D-Ala to D-Glu when Alr1 is active, the tight control of dat through translational coupling could prevent the depletion of the intracellular reserves of D-Ala necessary to overcome Ddl inhibition during acetate intoxication. Thus. the elevated D-Ala pool maintained within the cell could represent a strategic adaptation by S. aureus to combat Ddl inhibition caused by organic acids typically present in the niches colonized by this bacterium.

In conclusion, our findings demonstrate that Ddl is the primary target of acetate anion intoxication in S. aureus. However, other biologically relevant organic anions like lactate, propionate and itaconate could also inhibit the alr1 mutant similar to acetate. Furthermore, the growth inhibition of the alr1 mutant by these organic acids could be rescued following D-Ala supplementation, which suggests that Ddl is a bona fide and conserved target of various organic acid anions. Indeed, it is tempting to speculate that the robust Alr1 activity leading to the accumulation of millimolar levels of D-Ala may have evolved in part to offset the inhibition of Ddl from the toxic effects of organic anions.

eLife Assessment

In this valuable study, the authors use Staphylococcus aureus to understand how organic acids inhibit bacterial growth. They provide convincing evidence that acetic acid specifically inhibits the activity of the Ddl enzyme and that S. aureus maintains a high intracellular D-ala concentration to circumvent acetate-mediated growth inhibition. This work will be of interest to researchers studying bacteria and antimicrobials.

Reviewer #1 (Public Review):

Summary:

The manuscript entitled "Staphylococcus aureus counters organic acid anion-mediated inhibition of peptidoglycan cross-linking through robust alanine racemase activity" by Panda, S et al. reports an extensive biochemical analysis of the result from a Tn screen that identified alr1 as being required for acetic acid tolerance. In the end, they demonstrate that reduced D-Ala pools in the ∆alr1 mutant lead to a drastic reduction in D-Ala-D-Ala dipeptide. They show that this is due to the ability of organic acid anions to limit the D-Ala-D-Ala ligase enzyme Ddl. They demonstrate that:

(1) Acetate exposure in the ∆alr1 results in reduced D-Ala-D-Ala dipeptide, but not the monomers.

(2) Acetate can bind to purified Ddl in vitro.

(3) This binding results in reduced enzyme activity.

(4) Other organic acid anions such as lactate, proprionate, and itaconitate can also inhibit Ddl.

The experiments are clearly described and logically laid out. I have only a few minor comments to add.

Strengths:

The most significant strength is the exceptional experimental data that supports the authors' hypotheses.

Weaknesses:

Only minor weaknesses were identified by this reviewer.

(1) Which allele is alr1, the one upstream of MazEF or the one in the Lysine biosynthetic operon?

(2) Figure 3B. Where does the C3N2 species come from in the WT and why is it absent in the mutants? It is about 25% of the total dipeptide pool.

(3) Figure 3D could perhaps be omitted. I understand that the authors attained statistical significance in the fitness defect, but biologically this difference is very minor. One would have to look at the isotopomer distribution in the Dat overexpressing strain to make sure that increased flux actually occurred since there are other means of affecting activity (e.g. allosteric modulators).

(4) In Figure 4A, why is the complete subunit UDP-NAM-AEKAA increasing in each strain upon acetate challenge if there was such a stark reduction in D-Ala-D-Ala, particularly in the ∆alr1 mutant? For that matter, why are the levels of UDP-NAM-AEKAA in the ∆alr1 mutant identical to that of WT with/out acetate?

(5) Figure 4B. Is there no significant difference between ddl and murF transcripts between WT and ∆alr1 under acetate stress? This comparison was not labeled if the tests were done.

(6) Although tricky, it is possible to measure intracellular acetate. It might be of interest to know where in the Ddl inhibition curve the cells actually are.

Reviewer #2 (Public Review):

Summary:

In this manuscript, using Staphylococcus aureus as a model organism, Panda et al. aim to understand how organic acids inhibit bacterial growth. Through careful characterization and interdisciplinary collaboration, the authors present valuable evidence that acetic acid specifically inhibits the activity of Ddl enzyme that converts 2 D-alanine amino acids into D-ala-D-ala dipeptide, which is then used to generate the stem pentapeptide of peptidoglycan (PG) precursors in the cytoplasm. Thus, a high concentration of acetic acid weakens the cell wall by limiting PG-crosslinking (which requires a D-ala portion). However, S. aureus maintains a high intracellular D-ala concentration to circumvent acetate-mediated growth inhibition.

Strengths:

The authors utilized a well-established transposon mutant library to screen for mutants that struggle to grow in the presence of acetic acid. This screen allowed authors to identify that a strain lacking intact alr1, which encodes for alanine racemase (converts L-ala to D-ala), is unable to grow well in the presence of acetic acid. This phenotype is rescued by the addition of external D-ala. Next, the authors rule out the contribution of other pathways that could lead to the production of D-ala in the cell. Finally, by analyzing D-ala and D-ala-D-ala concentrations, as well as muropeptide intermediates accumulation in different mutants, the authors pinpoint Ddl as the specific target of acetic acid. In fact, the synthetic overexpression of ddl alone overcomes the toxic effects of acetic acid. Using genetics, biochemistry, and structural biology, the authors show that Ddl activity is specifically inhibited by acetic acid and likely by other biologically relevant organic acids. Interestingly, this mechanism is different from what has been reported for other organisms such as Escherichia coli (where methionine synthesis is affected). It remains to be seen if this mechanism is conserved in other organisms that are more closely related to S. aureus, such as Clostridioides difficile and Enterococcus faecalis.

Weaknesses:

Although the authors have conclusively shown that Ddl is the target of acetic acid, it appears that the acetic acid concentration used in the experiments may not truly reflect the concentration range S. aureus would experience in its environment. Moreover, Ddl is only significantly inhibited at a very high acetate concentration (>400 mM). Thus, additional experiments showing growth phenotypes at lower organic acid concentrations may be beneficial. Another aspect not adequately discussed is the presence of D-ala in the gut environment, which may be protective against acetate toxicity based on the model provided.

Author response:

The following is the authors’ response to the original reviews.

Public reviews:

Reviewer #1:

(1) Which allele is alr1, the one upstream of mazEF or the one in the lysine biosynthetic operon?

Alr1 is encoded by SAUSA300_2027 and is the gene upstream to mazEF. We have now incorporated this information in the manuscript (Line# 127).

(2) Figure 3B. Where does the C3N2 species come from in the WT and why is it absent in the mutants? It is about 25% of the total dipeptide pool.

In Figure 3B, C3N2 species results from the combination of C3N1 (from Alr1) and C0N1 (from Dat). The reason this species is completely absent in either of the two mutants is because it requires one D-Ala from both Alr1 and Dat proteins to generate C3N2 D-Ala-D-Ala.

(3) Figure 3D could perhaps be omitted. I understand that the authors attained statistical significance in the fitness defect, but biologically this difference is very minor. One would have to look at the isotopomer distribution in the Dat overexpressing strain to make sure that increased flux actually occurred since there are other means of affecting activity (e.g. allosteric modulators).

Thank you for the suggestion. We agree with the reviewer that the fitness defect observed after increased dat expression is relatively minor and have moved this figure to the supplementary section as Figure 3-figure supplement 1.

Although we attempted to amplify the fitness defect of dat expression by cloning dat on to a multicopy vector, we couldn't maintain its stable expression in S. aureus. This instability may be due to the depletion of D-Ala when dat is overexpressed. As a result, we switched to expressing dat from a single additional copy integrated into the SaPI locus, which was sufficient to cause the expected fitness defect, albeit a minor one.

(4) In Figure 4A, why is the complete subunit UDP-NAM-AEKAA increasing in each strain upon acetate challenge if there was such a stark reduction in D-Ala-D-Ala, particularly in the ∆alr1 mutant? For that matter, why are the levels of UDP-NAM-AEKAA in the ∆alr1 mutant identical to that of WT with/out acetate?

Thank you for raising this important point. We have addressed this in line# 299-302 and 451-455 of the revised manuscript. In short, we believe that the inhibition of Ddl by acetate significantly increases the intracellular pool of the tripeptide UDP-NAM-AEK, which then outcompetes the substrate (pentapeptide; UDP-NAM-AEKAA) of MraY. As a result, the intracellular concentration of the pentapeptide increases since it is no longer efficiently consumed by MraY. This explanation is also supported by a kinetic study conducted in Ref (1), where the competition between UDP-NAM-AEKAA and UDP-NAM-AEK as substrates for MraY is demonstrated.

(5) Figure 4B. Is there no significant difference between ddl and murF transcripts between WT and ∆alr1 under acetate stress? This comparison was not labeled if the tests were done.

Thank you for suggesting this comparison. The ddl and murF transcripts between WT and alr1 under acetate stress were significantly different. We have added this comparison to Figure 4B.

(6) Although tricky, it is possible to measure intracellular acetate. It might be of interest to know where in the Ddl inhibition curve the cells actually are.

Thank you for the suggestion. We agree this would have been an excellent addition to the manuscript. However, accurately measuring intracellular acetate would require the use of radiolabeled acetate (2), and we currently lack the expertise to do this experiment. However, since our study clearly shows that acetate-mediated growth impairment is due to Ddl inhibition, and the IC50 of acetate for Ddl is around 400 mM, we predict that the intracellular concentration must be close to or above this IC50 to observe the growth phenotypes we report.

Reviewer #2:

Although the authors have conclusively shown that Ddl is the target of acetic acid, it appears that the acetic acid concentration used in the experiments may not truly reflect the concentration range S. aureus would experience in its environment. Moreover, Ddl is only significantly inhibited at a very high acetate concentration (>400 mM). Thus, additional experiments showing growth phenotypes at lower organic acid concentrations may be beneficial.

Thank you for the suggestion. In response to the reviewer, we have measured growth at various acetate concentrations and demonstrate a concentration-dependent effect (Figure 1C).

We use 20 mM acetic acid in our study. In the gut, where S. aureus colonizes, acetate levels can reach up to 100 mM, so we believe our concentrations are physiologically relevant. When S. aureus encounters 20 mM acetate, the intracellular concentration can rise to 600 mM if the transmembrane pH gradient is 1.5 units, which is well above the ~400 mM IC50 we report for Ddl.

Another aspect not adequately discussed is the presence of D-ala in the gut environment, which may be protective against acetate toxicity based on the model provided.

Thank you for pointing this out. We agree that D-Ala from the gut microbiota could protect against acetate toxicity, and we’ve included this in the discussion. However, our study clearly indicates that S. aureus itself maintains high intracellular D-Ala levels through Alr1 activity which is sufficient to counter acetate anion intoxication.

Recommendation for the authors:

Reviewer #2:

Major Comments:

(1) In Line 85, authors indicate S. aureus may encounter a high concentration of ~100 mM acetic acid (extracellular?). Could the authors cite more (and recent) references indicating S. aureus encounters >100 mM acetic acid in the environment?

To the best of our knowledge, no studies have specifically examined whether S. aureus encounters high mM concentration of acetate in the gut. Line 85 was surmised from multiple studies: recent findings that S. aureus colonizes the gut (3, 4) and that the gut environment has high acetate levels (~100 mM) (5). In response to the reviewers request, more recent references supporting high acetate concentrations in the gut (6, 7) have been added in Line# 86.

(2) In Line 117, it is mentioned that S. aureus when grown in vitro at 20 mM acetic acid can accumulate ~600 mM acetic acid in the cytoplasm.

a. Does the intracellular concentration go up proportionally if grown in 100 mM acetic acid? Given the IC50 of acetic acid-mediated inhibition of Ddl is ~400 mM, I wonder how physiologically relevant this finding presented here is.

Thank you for the opportunity to explain this further. If S. aureus encounters a concentration of 100 mM acetate and its transmembrane pH gradient (pHin-pHout) is held at 1.5, the intracellular concentration of acetate could theoretically increase up to 3 M based on Ref (8). However, previous studies have shown that bacteria can lower the magnitude of transmembrane pH gradient by decreasing their intracellular pH to limit accumulation of anions within cells (9, 10).

Although our study shows that the IC50 of Ddl inhibition by acetate is relatively high (~400 mM), we believe it’s still relevant because just 20 mM of environmental acetate at a pH of 6.0 can raise the intracellular concentration of acetate to over 600 mM, which is well above the IC50 we report for Ddl. Moreover, since S. aureus may encounter high concentrations of acetate during gut colonization, we believe our findings are physiologically relevant.

b. Could the authors show concentration-dependent growth inhibition in alr::tn by titrating a range of acetic acid concentrations (for example 0, 0.5, 1, 5, 10, 20 mM)? Measuring intracellular acetate concentration may be beneficial as well.

Thank you for this question. We now provide data to support that acetate-mediated inhibition of the alr1 mutant is concentration-dependent (see Figure 1C).

c. It appears that there may be excess D-ala in the gut environment (PMIDs: 30559391; 35816159), which could counter the high acetate based on the model presented here. Could the authors clarify and/or include this information in the manuscript?

This is an excellent point, and we have now included it in the discussion (Line# 470-475). It is indeed possible that D-Ala produced by the gut microbiome may further enhance S. aureus resistance to organic acid anions, in addition to the inherent contribution of Alr1 activity.

(3) The following is not needed; however, it would be interesting if the authors could show that S. aureus cells grown in the presence of acetate are highly sensitive to cycloserine (which targets Alr and Ddl) compared to cells grown in the absence of acetate.

Thank you for the suggestion. We are currently studying D-cycloserine (DCS) resistance in S. aureus. Although we provide the data below for clarification, it is not included in the current manuscript as it is part of a separate study.

As the reviewer speculated, S. aureus is more susceptible to DCS when grown in the presence of acetate (see figure below). Normally, complete growth inhibition occurs at 32 µg/ml of DCS. However, with 20 mM acetic acid present, complete inhibition is achieved at just 8 µg/ml of DCS. Furthermore, the growth inhibition is completely rescued when externally supplemented with 5 mM D-Ala. We believe that DCS works synergistically with acetate to inhibit Ddl activity, and we are conducting additional studies to explore this further.

Minor Comments:

(1) Many commas are missing.

Missing commas are now incorporated.

(2) Line 77: disassociate --> dissociate

Corrected.

(3) Line 103: that --> which

Corrected.

(4) Lines 199-203: authors could have used gfp/luciferase reporter to test their hypotheses.

Thank you for the suggestion. Initially, we created GFP translational fusions for all the mutants mentioned in Line# 199-203. However, the fluorescence intensity was too low to test the hypothesis, as these were single-copy fusions inserted at the SaPI site of the S. aureus genome. Because of this limitation, we took advantage of the essentiality of D-Ala-D-Ala in S. aureus to report on various mutants instead of a fluorescent reporter. In hindsight, a LacZ reporter assay might have been equally effective.

(5) Line 339: It would be beneficial to introduce that Ddl has two independent ATP and D-ala binding sites.

We have now added that information (Line# 338-339).

(6) Is ddl an essential gene? If so, explicitly mention that.

Yes, ddl is an essential gene and we have now incorporated this information in Line 103.

(7) Line 354: shows a difference in density?

The use of the term “difference density” is a technical crystallographic term commonly used to connote density observed for ligands in X-ray crystal structures. In this case, it simply refers to the observed density that corresponds to the two acetate ions bound within the Ddl active site.

(8) Line 498: "Thus." Typo, change period to comma.

We have corrected as suggested in Line 496.

(9) Figure 1 legend says "was screen" instead of screened.

This is now corrected.

(10) Figure 1- Figure Supplement 1B: including data for alr2::tn dat::tn may ensure no redundancy (Lines 171-172). It is currently missing.

Thank you for the suggestion. We now include both alr2dat double mutant and the alr1alr2dat triple mutant in Figure 1 - Figure Supplement 1B. In addition we also show that the alr1alr2dat mutant is resuced by the addition of D-Ala in Figure 1 - Figure Supplement 1C. The mutant information is also added to Table S5.

(11) Figure 7: pentaglycine coming off of NAM is misleading. Remove untethered pentaglycine bridges.

We thank you for pointing this out. We have modified the figure in the manuscript as suggested by the reviewer.

(12) Are alr1/ddl cells (with limited 4-3 PG crosslink) less sensitive to vancomycin?

On the contrary, the alr1 mutant is slightly more sensitive to vancomycin compared to the wild-type strain (see Figure below). We believe this happens because the alr1 mutant incorporates less D-Ala-D-Ala into the peptidoglycan, reducing the number of targets for vancomycin. As a result, vancomycin may be able to saturate the available D-Ala-D-Ala targets on the cell wall at a lower concentration in the alr1 mutant than in the wild type strain, leading to increased sensitivity. We haven’t included this data in the manuscript as it is part of a separate study.

(13) Based on the structural studies, could the authors mutate the residues of Ddl involved in acetic acid binding, thereby making it resistant to acetic acid stress?

The residues that the acetate anion interacts with are located within the ATP-binding and D-Ala-binding sites of Ddl. Since these residues are essential for Ddl function, we are unable to mutate them.

(14) Microscopy to show the cell morphologies of wild-type and mutants exposed to acetic acid (and with D-ala supplementation) could be potentially interesting.

Thank you for the suggestion. We did perform microscopy, expecting changes in cell shape or size, but the results were unremarkable and not included in the manuscript.

References:

(1) Hammes WP & Neuhaus FC (1974) On the specificity of phospho-N-acetylmuramyl-pentapeptide translocase. The peptide subunit of uridine diphosphate-N-actylmuramyl-pentapeptide. J Biol Chem 249(10):3140-3150.

(2) Roe AJ, McLaggan D, Davidson I, O'Byrne C, & Booth IR (1998) Perturbation of anion balance during inhibition of growth of Escherichia coli by weak acids. J Bacteriol 180(4):767-772.

(3) Acton DS, Plat-Sinnige MJ, van Wamel W, de Groot N, & van Belkum A (2009) Intestinal carriage of Staphylococcus aureus: how does its frequency compare with that of nasal carriage and what is its clinical impact? Eur J Clin Microbiol Infect Dis 28(2):115-127.

(4) Piewngam P_, et al._ (2023) Probiotic for pathogen-specific Staphylococcus aureus decolonisation in Thailand: a phase 2, double-blind, randomised, placebo-controlled trial. Lancet Microbe 4(2):e75-e83.

(5) Cummings JH, Pomare EW, Branch WJ, Naylor CP, & Macfarlane GT (1987) Short chain fatty acids in human large intestine, portal, hepatic and venous blood. Gut 28(10):1221-1227.

(6) Correa-Oliveira R, Fachi JL, Vieira A, Sato FT, & Vinolo MA (2016) Regulation of immune cell function by short-chain fatty acids. Clin Transl Immunology 5(4):e73.

(7) Hosmer J, McEwan AG, & Kappler U (2024) Bacterial acetate metabolism and its influence on human epithelia. Emerg Top Life Sci 8(1):1-13.

(8) Carpenter CE & Broadbent JR (2009) External concentration of organic acid anions and pH: key independent variables for studying how organic acids inhibit growth of bacteria in mildly acidic foods. J Food Sci 74(1):R12-15.

(9) Russell JB (1992) Another explanation for the toxicity of fermentation acids at low pH: anion accumulation versus uncoupling. Journal of Applied Bacteriology 73(5):363-370.

(10) Russell JB & Diez-Gonzalez F (1998) The effects of fermentation acids on bacterial growth. Adv Microb Physiol 39:205-234.

Version 2

Abstract

Abstract

Weak organic acids are commonly found in host niches colonized by bacteria, and they can inhibit bacterial growth as the environment becomes acidic. This inhibition is often attributed to the toxicity resulting from the accumulation of high concentrations of organic anions in the cytosol, which disrupts cellular homeostasis. However, the precise cellular targets that organic anions poison and the mechanisms used to counter organic anion intoxication in bacteria have not been elucidated. Here, we utilize acetic acid, a weak organic acid abundantly found in the gut to investigate its impact on the growth of Staphylococcus aureus. We demonstrate that acetate anions bind to and inhibit D-alanyl-D-alanine ligase (Ddl) activity in S. aureus. Ddl inhibition reduces intracellular D-alanyl-D-alanine (D-Ala-D-Ala) levels, compromising staphylococcal peptidoglycan cross-linking and cell wall integrity. To overcome the effects of acetate-mediated Ddl inhibition, S. aureus maintains a substantial intracellular D-Ala pool through alanine racemase (Alr1) activity and additionally limits the flux of D-Ala to D-glutamate by controlling D-alanine aminotransferase (Dat) activity. Surprisingly, the modus operandi of acetate intoxication in S. aureus is common to multiple biologically relevant weak organic acids indicating that Ddl is a conserved target of small organic anions. These findings suggest that S. aureus may have evolved to maintain high intracellular D-Ala concentrations, partly to counter organic anion intoxication.

Introduction

Organic acids produced by host and bacterial metabolism are critical determinants of infection outcomes (1, 2). During infection, the host macrophages produce millimolar amounts of itaconate, a dicarboxylic acid known to inhibit bacterial growth (3). Conversely, many bacterial pathogens and the gut microflora excrete short-chain organic fatty acids, which exhibit immunomodulatory functions and can skew the host response during infection (4, 5). Upon entry into the bacterial cell, organic acids can become toxic to bacteria when they dissociate in the cytosol as protons and organic anions. The influx of protons can result in cytoplasmic acidification and prove lethal for some pathogens if not adequately controlled (6). Similarly, organic anions have been shown to accumulate to toxic levels in the bacterial cytoplasm (7). However, the precise consequences of organic anion toxicity and the mechanisms pathogens employ to withstand the effects of anion perturbations within cells are not clearly understood.

Here, we focus on the response of Staphylococcus aureus to acetic acid, which is the primary end-product of glucose catabolism under aerobic conditions. S. aureus also likely encounters high concentrations (up to 100 mM) of acetic acid and other short-chain fatty acids produced by human gut microbiota during intestinal colonization (8–10). On average, 20% of adults carry S. aureus in their intestines (11), and the burden there often surpasses that found in nasal passages by more than three orders of magnitude, establishing the gut as a primary site for S. aureus colonization (12). We have previously shown that excreted acetic acid can promote cytoplasmic acidification in cultures of S. aureus, especially when the external environment becomes sufficiently acidic (pH< 5) (13). Cytoplasmic acidification promotes protein oxidation and triggers a staphylococcal ClpP-dependent damage response that eliminates unfit cells from the population (14). In contrast, in mildly acidic environments (pH 5.5-6.5), although S. aureus actively buffers its intracellular environment against acidification, the transmembrane pH gradient (ΔpH) of S. aureus will drive the accumulation of millimolar quantities of acetate anions into the cytoplasm. Previous studies in Escherichia coli have shown that acetate intoxication causes an osmotic imbalance that can transiently be accommodated by the efflux of physiological anions like glutamate (7). In addition, acetate anions have also been reported to impact enzymes in the methionine biosynthetic pathway, resulting in a toxic accumulation of homocysteine and a reduction in intracellular methionine leading to growth inhibition of E. coli (15, 16). However, it remains unclear if these effects are common to other bacteria.

Here, we demonstrate that the primary target of acetate intoxication in S. aureus is Ddl. Ddl is essential for staphylococcal growth and produces D-Ala-D-Ala dipeptide which is incorporated into peptidoglycan to cross-link the peptide side chains of neighboring glycan strands. We also demonstrate that carbon flux through alanine racemase and a tight control of Dat activity increases the cytosolic D-Ala pool to counter acetate-mediated inhibition of Ddl. Importantly, these phenotypes are not unique to acetate, but are conserved across multiple biologically important organic acid anions. Therefore, we propose that S. aureus may have evolved to maintain a high intracellular D-Ala pool, partly to offset the inhibition of Ddl by organic anions typically encountered during human colonization.

Results

Alanine racemase counters acetate intoxication

To identify genetic determinants that counter the effects of acetate intoxication, we screened the Nebraska Transposon Mutant Library (NTML) for mutants sensitive to 20 mM acetic acid in Tryptic Soy Broth (TSB) media, pH 6.0. Under these conditions, S. aureus maintains its intracellular pH approx. 1.5 units above the external pH (17) and is estimated to accumulate over 600 mM acetate in the cytosol (18). The NTML strains were grown under static conditions at 37°C, and the extent of growth was determined at 24 h by measuring the optical density at 600 nm (OD600). As a control, we performed an identical screen without acetic acid supplementation. We normalized the growth of each mutant in both screens (± acetic acid) to their isogenic wild-type (WT) strain. A comparison of growth indices (OD600 Tn-mut/WT) for each mutant in the presence and absence of 20 mM acetic acid revealed that most mutants clustered close to an index of 1 in the plot (Figure 1A), which suggested that most mutants tolerated acetate intoxication reasonably well. A few mutants that grew poorly following acetate intoxication due to inherent growth defects were observed close to the plot diagonal, whereas those mutants that did not have intrinsic growth deficiencies were located further away from the diagonal. Among the latter class of mutants, we observed that the alr1 mutant (SAUSA300_2027) had the most substantial reduction in growth when subjected to acetate stress (Figure 1A, B), with the severity of the phenotype depending on the acetate concentration (Figure 1C). To confirm that the acetate-dependent growth defect of the alr1 mutant was not due to polar effects, we complemented the mutant by inserting a functional copy of alr1 under the control of its native promoter into the S. aureus pathogenicity island (SaPI) attachment site. Genetic complementation completely restored the alr1 mutant phenotype to WT levels (Figure 1B, C). These results suggest that acetate intoxication impairs the growth of S. aureus in the absence of a functional alanine racemase. Further supporting this conclusion, we could reduce acetate toxicity in the alr1 mutant by culturing this strain in glucose- free TSB media, which alleviates carbon catabolite repression and activates TCA cycle- dependent acetate catabolism (Figure 1D) (19). Conversely, the inactivation of citrate synthase (citZ), the first enzyme of the TCA-cycle, re-imposed acetate toxicity in the alr1 mutant when cultured in glucose-free TSB media (Figure 1D).

Alanine racemase activity counters acetate intoxication

(A) The Nebraska Transposon Mutant library was screened against 20 mM acetic acid, pH 6.0 to identify mutants with altered growth phenotypes. The WT strain and transposon mutants were grown for 24 h in TSB ± 20 mM acetic acid. The bacterial growth at 24 h was measured spectrophotometrically (OD600) and normalized to WT growth. The X and Y-axis on the plot represent normalized growth values for each mutant in the presence or absence of acetate. (B) The growth of the WT, alr1 mutant, and alr1 complemented strain (alr1C) in TSB supplemented with 20 mM acetic acid. (C) The tolerance of strains to various acetate concentrations was assessed by monitoring growth (OD600) following a previously published method (58). To maintain a transmembrane pH gradient of ∼1.5, the culture media was adjusted to pH 6, prior to challenging with subinhibitory acetate concentrations (40 mM-1.25 mM; two-fold dilutions). The relative growth (fractional area) of both the WT, alr1 and alr1C mutant was calculated by comparing the area under the growth curves at subinhibitory concentrations of acetate to their corresponding controls (no acetate) and plotted against acetate concentrations. (D) Aerobic growth of WT, alr1, citZ, citZalr1 mutants in TSB media lacking glucose, but supplemented with 20 mM acetic acid. (E) LC-MS/MS analysis was performed to quantify the intracellular D-Ala-D-Ala pool in strains cultured for 3 h (exponential phase) in TSB ± 20 mM acetic acid. (F) The growth of strains was monitored following D-Ala supplementation (5 mM) in TSB + 20 mM acetate, (n=3, mean ± SD). Ac, acetate. ***, P value <0.001; ****, P value <0.0001.

Acetate intoxication alters the intracellular D-Ala-D-Ala pools

Alr1 catalyzes the conversion of L-Ala to D-Ala during staphylococcal growth (Figure 1-figure supplement 1A). The D-Ala is further converted to D-Ala-D-Ala dipeptide by the ATP-dependent Ddl (Figure 1-figure supplement 1A) and incorporated into peptidoglycan (PG) muropeptide, thus playing a crucial role in PG biosynthesis, cross-linking, and integrity (20, 21). Therefore, we hypothesized that under acetate stress, low concentrations of D-Ala in the alr1 mutant might concomitantly reduce D-Ala-D-Ala concentrations in the cell resulting in a growth defect. To test this hypothesis, we determined the intracellular pool of D-Ala-D-Ala using liquid chromatography- tandem mass spectrometry (LC-MS/MS). In regular growth media (TSB), we observed that the inactivation of alr1 decreased the D-Ala-D-Ala pool by approximately 80% compared to the WT strain (Figure 1E). However, following acetate intoxication, the level of D-Ala-D-Ala was depleted by more than 99% (Figure 1E). The external supplementation of D-Ala (5 mM) in the media fully restored the growth of the alr1 mutant to WT levels under acetic acid stress (Figure 1F), which suggests that increased intracellular D-Ala pools can overcome the detrimental impact of acetate intoxication.

The depletion of D-Ala-D-Ala following acetate intoxication is surprising since S. aureus is predicted to have two additional pathways that can synthesize D-Ala and channel it to the production of this dipeptide. For instance, S. aureus harbors a second predicted alanine racemase (Alr2) that could compensate for the lack of Alr1 activity (Figure 1-figure supplement 1A). Alternatively, Dat, which catalyzes the formation of D-Ala from pyruvate and D-glutamate (D- Glu), may functionally complement the alr1 mutant under acetate stress (Figure 1-figure supplement 1A). However, the lack of functional complementation from these alternate pathways of D-Ala biosynthesis following acetate intoxication suggests that not all metabolic routes to D- Ala are operational, or that regulatory bottlenecks limit pathway activity. To test these possibilities, we constructed a series of mutants in which all three predicted routes of D-Ala biosynthesis (alr1, alr2 and dat) were disrupted either individually or in various combinations and performed growth assays (Figure 1-figure supplement 1B). Surprisingly, we observed that the inactivation of alr1 and dat simultaneously (alr1dat mutant) was synthetic lethal in S. aureus, suggesting that alr1 and dat were the sole contributors of D-Ala in S. aureus. Indeed, the supplementation of D-Ala fully restored the growth of the alr1dat mutant (Figure 1-figure supplement 1C).

The inactivation of alr2, either alone or in combination with other D-alanine-generating enzymes, did not affect growth (Figure 1-figure supplement 1B). This observation suggests that alr2 is unlikely to be a functional alanine racemase under the growth conditions tested. Collectively, these results indicate that Dat activity accounts for D-Ala production in the absence of alr1, but its contribution is insufficient to counter acetate intoxication.

Insufficient translation of dat impacts the alr1 mutant following acetate intoxication

Since Dat activity contributes to D-Ala production in the alr1 mutant, we questioned why Dat is insufficient to sustain D-Ala-D-Ala pools under conditions of acetate intoxication. One possible explanation may relate to the maintenance of osmotic balance by S. aureus. It has been proposed that the intracellular accumulation of acetate anions may bring about an efflux of L/D- Glu from cells to adjust for osmolarity, thus exhausting one of the key substrates for Dat activity and limiting D-Ala production (7). However, this hypothesis is improbable since the expression of dat from a multicopy vector rescued the alr1 mutant from the effects of acetate intoxication (Figure 2-figure supplement 1A), suggesting that the intracellular D-Glu pools are sufficient to support D-Ala production through Dat activity. Alternatively, we hypothesized that the alr1 mutant’s heightened sensitivity to acetate toxicity could be due to a decrease in dat transcription which would effectively reduce intracellular D-Ala. However, we found no detrimental effect of acetate intoxication on dat transcription in the alr1 mutant (Figure 2-figure supplement 1B). Together, these observations raise the possibility that the depletion of D-Ala-D-Ala in the alr1 mutant following acetate intoxication may arise from a post-transcriptional regulatory bottleneck, that limits dat from meeting the demand for intracellular D-Ala.

In S. aureus, dat is part of a bicistronic operon (Figure 2A). The first gene, pepV, encodes an extracellular dipeptidase (22, 23). Transcriptional start site (TSS) mapping of the pepV-dat operon by the adaptor and radioactivity-free (ARF-TSS) method revealed a 30-nucleotide untranslated region (5’-UTR) extending upstream from the pepV initiation codon. The 5’-UTR includes a Shine-Dalgarno motif (ribosome binding site, SD1) upstream of the pepV start codon (Figure 2A). In addition, a second SD motif (SD2) associated with dat was identified within the pepV coding region (Figure 2A), and did not overlap with the pepV termination codon. The location of SD2 within pepV suggests that the insufficient production of D-Ala by dat following acetate intoxication could be attributed to suboptimal translation of dat. This could occur as ribosomes (70S) that are moving from SD1 may interfere with the translation of dat from SD2. To test this hypothesis, we engineered a nonsense mutation in pepV (alr1pepVQ12STOP mutant) that would prevent the ribosomes originating from SD1 from moving forward (Figure 2A). However, the alr1pepVQ12STOP mutant grew poorly compared to the alr1 mutant following acetate intoxication (Figure 2B). This suggested that the translation of dat is coupled to that of pepV presumably through stable mRNA secondary structures that form within pepV. These structures may not be effectively resolved in the alr1pepVQ12STOP mutant due to the absence of ribosome traffic on pepV mRNA.

Translational coupling of dat to pepV limits the alr1 mutant from countering acetate intoxication

(A) Schematic representation of various engineered mutations in the pepV- dat locus.SD, Shine-Dalgarno motif; TSS, transcriptional start site. (B)-(D) Growth of engineered mutants was monitored spectrophotometrically (OD600) in TSB supplemented with 20 mM acetate (n=3, mean ± SD). (E) RT-qPCR to determine dat transcription in various mutants relative to the WT strain.

As an alternative approach to determine if SD2 positioning within pepV impeded dat translation, we deleted pepV along with SD1 in the alr1 mutant (alr1pepVΔSD1-467, Figure 2A). In the resulting strain, dat translation was under the sole control of its native SD2. Remarkably, the alr1pepVΔSD1-467 mutant did not display a heightened sensitivity to acetate stress and grew identical to the WT strain following acetate intoxication (Figure 2C). Similarly, an alr1 mutant in which dat was linked to SD1 (alr1pepV mutant, Figure 2A) also phenocopied the WT strain following acetate intoxication (Figure 2D). Notably, the observed growth differences in alr1pepVΔSD1-467, alr1pepV and alr1pepVQ12STOPmutants following acetate intoxication did not result from any changes in dat transcription (Figure 2E). These findings collectively suggest that the native promoter elements, as well as the SD sites of pepV and dat can independently support the robust expression and translation of dat to levels required for countering acetate intoxication. However, the genetic arrangement of the dat translation initiation region (TIR) within pepV, offered tight control of dat translation, and prevented cells from producing sufficient enzyme following acetate intoxication.

Why is the Dat tightly controlled?

The need to tightly control Dat activity suggests that flux between D-Ala and D-Glu pools must be carefully balanced during staphylococcal growth. To gain insight into this process, we profiled the mass isotopologue distribution (MID) of D-Ala-D-Ala in the WT, alr1, and dat mutants under isotopic steady-state conditions using 13C 15N -L-Ala as the tracer during growth experiments in chemically defined medium (CDM). The flux of 13C315N1-L-Ala through Alr1 should result in 13C 15N -D-Ala production (Figure 3A, D-Ala retains labeled nitrogen). On the other hand, staphylococcal alanine dehydrogenases (Ald1 and Ald2) catalyze the conversion of 13C 15N -L- Ala to 13C3-pyruvate, and finally 13C3-D-Ala through Dat activity (Figure 3A). Thus, the labeled nitrogen in 13C 15N -L-Ala is lost as 15N -NH when fluxed through the Ald/Dat pathway (Figure 3A). Since the intracellular pools of D-Ala are converted to D-Ala-D-Ala, the MID of the latter metabolite should mirror the isotopologue ratios of D-Ala produced from either Alr1 or Dat activities.

Reaction orientation and fluxes through Alr1 and Dat

(A) Schematic representation of various isotopologues of D-Ala-D-Ala and D-Glu generated from 13C315N1 labeled L-Ala. Metabolites in blue mainly arise from Alr1, red, through the Ald1/2-Dat pathway and yellow are unlabeled intermediates within cells. The mass isotopologue distribution of (B) D-Ala-D-Ala and (C) D-Glu were determined by LC-MS/MS following the growth of S. aureus in chemically defined media supplemented with 13C315N1 L-Ala (n=3, mean ± SD). Isotopologues of D-Ala-D-Ala shown in grey color are minor species and are noted in Table S9.

LC-MS/MS analysis revealed that ∼80% of the intracellular D-Ala-D-Ala pool had incorporated the labeled L-Ala supplemented in media (fractional contribution, 0.80). As expected, the majority (∼55%) of the D-Ala-D-Ala in the WT was composed of the C6N2 isotopologue (in which both units of D-Ala contain labeled carbon and nitrogen), which suggested that alr1 was the major contributor of D-Ala in S. aureus (Figure 3B). Surprisingly, the sole contribution of dat activity (C6N0, C3N0, C0N0) to D-Ala-D-Ala was less than 1% in the WT strain, and D-Ala-D-Ala isotopologues with at least one D-Ala originating from dat activity (C6N1, C3N1, C3N2) although readily observed, were still in the minority. However, the D-Ala-D-Ala originating from Dat activity expanded substantially upon alr1 mutation (Figure 3B). Inactivation of dat itself displayed few differences in the MID of D-Ala-D-Ala, compared to the WT strain (Figure 3B). These results suggest that flux through Dat is most likely driven towards D-Glu in the WT strain rather than D- Ala. Only upon inactivation of alr1 does the Dat activity reverse towards the production of D-Ala. To confirm these predictions, we measured the levels of 15N1-D-Glu in the WT, alr1, and dat mutants following growth with the 13C315N1-L-Ala tracer. Consistent with Dat activity funneling D-Ala to D-Glu in the WT, approximately 78% of the D-Glu pool in the WT strain was 15N labeled. Furthermore, we observed that inactivation of dat resulted in the complete depletion of intracellular levels of 15N1-D-Glu (Figure 3C). Inactivation of alr1 also had a similar outcome with loss of 15N1-D-Glu pools due to the lack of 13C 15N1-D-Ala in this mutant (Figure 3C). Together, these results strongly suggest that in the WT strain, Dat activity diverts D-Ala towards D-Glu production.

Given the critical need to produce D-Ala-D-Ala during acetate intoxication, any diversion of its precursor pool (D-Ala) to D-Glu through Dat activity is bound to decrease cell fitness and thus may justify its tight translational control. To test this hypothesis, we determined the mean competitive fitness (w) of cells that overexpressed dat compared to those that had native levels of expression. Accordingly, we performed coculture competition assays of the WT strain with an isogenic mutant strain that either harbored an empty vector (p) integrated into the SaPI chromosomal site or a vector containing dat under control of its native promoter (pAS8), following acetate intoxication. Consistent with increased Dat activity in the WT strain being detrimental to the cell, the mean competitive fitness of the dat overexpressing strain was significantly lower (w4h= 0.91) in the exponential growth phase than its isogenic WT strain that harbored the empty vector (w4h= 1.26) (Figure 3D). Collectively, these results suggest that Dat catalyzes the production of D-Glu in the WT strain, and its tight regulation prevents excessive flux of D-Ala to D-Glu which is necessary to maintain cell fitness following acetate intoxication.

Acetate intoxication impacts PG biosynthesis

Since acetate intoxication ultimately affects D-Ala-D-Ala pools (Figure 1D), we predicted potential alterations to PG biosynthesis and cell wall integrity. To test this hypothesis, we quantified various cytosolic PG intermediates in the WT strain by LC-MS/MS analysis. Acetate intoxication caused a significant increase in the intracellular pools of multiple PG biosynthetic intermediates, including Uridine diphosphate N-acetylglucosamine (UDP-NAG), UDP-N- acetylmuramic acid (UDP-NAM), UDP-NAM-L-Ala, UDP-NAM-L-Ala-D-Glu-L-Lys and UDP-NAM-L-Ala-D-Glu-L-Lys-D-Ala-D-Ala (UDP-NAM-AEKAA) in the WT strain when compared to the unchallenged control (Figure 4A). However, the growth of the WT strain was slightly inhibited by acetic acid (Figure 1B), which suggests that the observed accumulation of PG intermediates may have been due to an imbalance between the rates of PG biosynthesis and growth. Notably, the alr1 mutant showed higher levels of UDP-NAM-AEK compared to the WT and the dat mutant following acetate intoxication (Figure 4A), indicating a metabolic block in the production of UDP- NAM-AEKAA due to insufficient D-Ala-D-Ala. The effect of this metabolic block is also evident from the increased transcription of ddl and murF (Figure 4B) which encode enzymes that incorporate D-Ala-D-Ala into PG precursors, suggesting a greater need to maintain peptidoglycan cross-linking following acetate intoxication.

Acetate intoxication impacts soluble PG precursor pools and cell wall cross- linking.

(A) The intracellular pool of PG intermediates in exponential phase cultures of S. aureus was estimated using LC-MS/MS analysis. cps, counts per second (B) ddl and murF transcription in the exponential growth phase was determined by RT-qPCR analysis (n=3, mean ± SD). (C) Cell wall muropeptide analysis of the WT, alr1 and dat mutants was determined following growth in TSB ± 20 mM acetate for 3 h. Cell wall cross-linking was estimated as previously described (53). Ac, acetate. *, P value <0.05; **, P value <0.01; ***, P value <0.001; ****, P value <0.0001.

Unsurprisingly, the dysregulation of D-Ala-D-Ala homeostasis following acetate intoxication was also reflected in the extent of cell wall cross-linking in the WT, alr1 and dat mutants. Muropeptide analysis revealed that acetate intoxication in the WT strain increased levels of monomeric muropeptides (Figure 4C, Figure 4-figure supplement 1). Conversely, the percentage of di- and trimeric muropeptides decreased relative to the WT control, as did the percent cross- linking (Figure 4C). These observations suggest that acetate intoxication constrains the D-Ala-D- Ala pool in the WT strain and alters PG cross-linking despite Alr1 activity. The extent of PG cross- linking in the dat mutant was similar to WT in the presence or absence of acetate, consistent with our finding that the Dat activity plays a limited role in maintaining the D-Ala-D-Ala pool in the WT strain (Figure 4C). In contrast, PG cross-linking in the alr1 mutant was lower than the WT strain by ∼10% (Figure 4C). Acetate intoxication further decreased the cross-linking approximately 20% relative to WT as well as the ratio of dimeric to monomeric muropeptides in the alr1 mutant, which inevitably reduced the growth of this strain (Figure 4C).

Muropeptide analysis also revealed the accumulation of a disaccharide tripeptide (NAG- NAM-AEK (M3); m/z. Da, 826.4080) in the peptidoglycan (PG) extracted from the alr1 mutant (Figure 4-figure supplement 1B). This finding suggests that the significantly elevated levels of UDP-NAM-AEK in the alr1 mutant could efficiently outcompete the substrate specificity of phospho-N-acetylmuramyl pentapeptide translocase (MraY) for UDP-NAM-AEKAA, ultimately becoming integrated into the PG structure itself. Interestingly the incorporation of UDP-NAG- NAM-AEK into the alr1 mutant’s PG only marginally increased following acetate treatment (Figure 4-figure supplement 1B, see inset). The increase of UDP-NAG-NAM-AEK is most likely an underestimate since cells with higher levels of incorporation are more likely to lyse due to a reduction in PG cross-linking. Overall, these observations support a model wherein the immediate consequences of acetate intoxication are defects in PG crosslinking and biosynthesis.

Acetate intoxication inhibits Ddl activity

While the above observations point to the consequences of acetate intoxication of S. aureus, its molecular target was not initially identified. Since acetate intoxication dramatically reduces D- Ala-D-Ala levels in the alr1 mutant (Figure 1D), we reasoned that acetate might inhibit either Dat or Ddl activity. To distinguish between these two targets, we measured the levels of D-Ala in the alr1 mutant following acetate intoxication. Surprisingly, we observed that the D-Ala pools in the alr1 mutant did not significantly change in response to acetate intoxication compared to the untreated control (Figure 5A). This suggested that Dat activity was preserved in the alr1 mutant to the same extent as its untreated control and was not affected by acetate.

Acetate anion inhibits Ddl activity.

(A) The intracellular D-Ala was determined by LC- MS/MS analysis. (B) The ddl gene was overexpressed in S. aureus using a cadmium inducible expression system (pSP36). CdCl2, 0.312 µM. (C) Inhibition of recombinant His-tagged Ddl activity in the presence of 300 mM sodium acetate (D) IC50 curve of the inhibition of rDdl by acetate. Michaelis-Menten kinetics of rDdl in varying concentrations of (E) D-Ala, and (F) ATP in the presence of acetate to assess the inhibition mechanism. (G) Structure of the acetate bound Ddl (PDB:8FFF). (H) Acetate bound to the ATP binding site of Ddl (I) Acetate bound to the second D-Ala binding site of Ddl. The calculated Fo-Fc omit maps are contoured to 3σ and the mesh is shown in blue. (J) Superimposed structure of acetate bound Ddl (slate blue) with StaDdl apo structure (PDB:2I87, beige) and StaDdl-ADP complex structure (PDB:2I8C, grey) showing a shift of ω loop (red) to ATP binding site. The D-Ala-D-Ala was modeled at the D-Ala binding site using Thermos thermophius HB8 Ddl structure (PDB:2ZDQ). The bound ADP (grey) of PDB:2I87 and modeled D-Ala-D-Ala (light blue) indicates the positioning of Ac at ATP and second D-Ala binding sites respectively. Ac, acetate; V, velocity; *, P value <0.05; ****, P value <0.0001.

Conversely, these findings also indicate that the acetate-dependent decrease of the D-Ala-D- Ala pool in the alr1 mutant was most likely due to the inhibition of Ddl. To test this hypothesis, we cloned S. aureus ddl under the control of a cadmium inducible promoter and induced its expression in the alr1 mutant following acetate intoxication (Figure 5B). Indeed, the growth of the alr1 mutant was restored to WT levels when ddl was overexpressed, strongly suggesting that Ddl was the target of acetate anion (Figure 5B).

To confirm that acetate inhibits Ddl through direct interactions, we undertook two separate approaches. As the first approach, 6xHis-tagged S. aureus Ddl was purified, and in-vitro enzyme kinetic assays were performed to determine the possible inhibitory mechanism of Ddl by acetate.

Considering the high concentration of acetate estimated to accumulate in the cytoplasm, a concentration of 300 mM sodium acetate was used in the initial reactions to test inhibition (Figure 5C). Interestingly, variation of acetate concentration showed that Ddl was inhibited in vitro, and these conditions suggest an IC50 of 400.3 ± 8 mM (Figure 5D). This indicates significant inhibition of Ddl by acetate when the cellular concentration is near the hypothesized 600 mM, further confirming that Ddl is a direct target of inhibition by acetate anion. Furthermore, based on kinetic experiments performed under varying concentrations of either ATP or D-Ala, the kcat values are shown to be distinctly different for each acetate concentration, which strongly suggests a mixed inhibition mechanism for acetate (Figure 5E and F, Table S1).

Differential Scanning Fluorometry (DSF) was used as another approach to assess the direct binding of acetate to Ddl (Table S2). Ddl has two D-Ala binding sites and one ATP-binding site (24), and DSF experiments were conducted with various combinations of these ligands. The Ddl protein without any ligand bound shows a melting temperature (Tm) of 45 °C. After adding 300 mM sodium acetate, Ddl exhibited a 3.7 °C Tm shift indicating a slight thermal stabilization upon binding acetate. This is higher than the shift in the Tm exhibited by a Ddl/ATP complex. The addition of ADP to Ddl results in a decrease of 3.2 °C, indicating a decrease in thermal stability compared to Ddl alone. Intriguingly, when adding acetate to Ddl complexes with ATP or ADP, the Tm increased to 48.9 °C and 49.9 °C, respectively (Table S2). This represents a Tm increase of 2.3 °C when acetate is added to a Ddl/ATP complex, but a Tm increase of 8.1 °C when acetate is added to a Ddl/ADP complex. The addition of D-Ala to the reaction mixture increases the Tm of Ddl by 4.2 °C and adding acetate to the Ddl/D-Ala mixture shows only a 0.3 °C Tm shift (Table S2). The widely varying changes in Tm for the tested complexes, particularly when comparing the Tm values for ligand-free, ADP-bound, and the ADP/Acetate complex, further support a mixed inhibition mechanism as these data suggest acetate may bind to multiple sites on Ddl or the location of these binding sites may change depending on the ligand-bound state of the enzyme due to Ddl conformational changes as observed in Ddl orthologs (24).

Ddl/Acetate complex structure shows binding of acetate at both substrate binding sites

To gain further insight into the mechanism of acetate inhibition, the X-ray crystal structure of a Ddl/acetate complex was obtained using co-crystals of Ddl and acetate. The crystal diffracted to 1.9 Å and data were consistent with a P 2 21 21 space group possessing one molecule of Ddl in the asymmetric unit (Table S3). The crystal structure of the Ddl/acetate complex (PDB:8FFF) shows difference density corresponding to acetate at two different sites of the protein. One acetate is positioned within the adenine binding subsite of the ATP binding site and the other acetate ion is positioned in the second D-Ala binding site (Figure 5G). The acetate ion in the ATP binding site interacts with the side chain of Lys177 and the backbone nitrogen of Val216. Also, the methyl group of acetate forms van der Waals interactions with the side chain of Leu145 (Figure 5H). The acetate ion that binds to the D-Ala binding site forms a bidentate polar interaction with the side chain of Arg291 and a hydrogen-bonded interaction with the backbone nitrogen of Gly312 (Figure 5I). These two residues are conserved in Ddl homologs and previous structural data clearly illustrate the crucial role these residues play in D-Ala binding (24).

The acetate-bound structure shows conformational differences compared to the previously published ligand-free and ADP-bound structures (24). The ω loop, which is associated with substrate binding, is disordered in both the S. aureus Ddl ligand-free (PDB:2I87) and the Ddl- ADP complex structures (PDB:2I8C) as well as other available Ddl crystal structures that lack bound substrates or ligands (PDB:3K3P, 5DMX and 6U1C) (25–27). Interestingly, this loop is well ordered in the acetate-bound structure described here (Figure 5J), which gives the first view of the S. aureus Ddl ω loop and the interactions it may form with substrates or inhibitors. The structural stabilization of the ω loop is consistent with the DSF results exhibiting an increase in the melting temperature upon binding acetate. The ω loop is shifted towards the ATP binding site and repositions the conserved Tyr246 side chain within the ATP binding site, which likely hinders the binding of ATP (Figure 5J). This positioning is comparable with the Mycobacterium tuberculosis Ddl (PDB:3LWB) ligand-free structure, which also takes a closed conformation showing the ω loop positioned within the ATP binding site and obstructing ATP binding (28). Taken together, the kinetic, DSF, and structural data suggest that while acetate can directly bind within both substrate binding pockets of Ddl, it also stimulates conformational changes in the dynamic ω loop to afford more allosteric-like effects on enzyme activity. Each of these observations support a mixed inhibition modality.

Multiple organic acids inhibit the alr1 mutant in a D-Ala-dependent manner

Finally, we determined whether the growth inhibition of the alr1 mutant is unique to acetate anion or is a more general phenomenon mirrored by addition of other small organic acids. Accordingly, we initially performed molecular docking studies of three biologically relevant organic anions: lactate, propionate and itaconate, in both the ATP and D-Ala binding pockets of Ddl (Figure 6A-D). The acetate anion-bound structure of Ddl was used as a reference for analysis. The docking results suggest reasonable poses for lactate, propionate and itaconate within the ATP binding site forming polar interactions with Ddl residues conserved for binding ATP. Upon docking, the carboxylate moieties of both lactate and propionate form ionic interactions with the Lys177 side chain similar to those observed in the Ddl/acetate crystal structure (Figure 6A and B). Also, the side chain of Glu213 in Ddl forms a hydrogen bonded interaction with the hydroxyl of lactate (Figure 6A) and van der Waals interactions between propionate and nearby side chains of Phe175 and Phe295 (Figure 6B) were indicated. The two carboxylate groups of itaconate form hydrogen bonded interactions with backbone amide nitrogen atoms of Ala218 and Tyr246 as well as a van der Waals interaction with the nearby side chain of Phe175 (Figure 6C).

Biologically relevant weak acids inhibit growth of the alr1 mutant.

Molecular docking of (A) lactate (B) propionate and (C) itaconate to the ATP binding site of Ddl. (D) The relative positions and poise of different organic anions in relation to acetate in the D-Ala binding site of Ddl was determined using Schrödinger Glide. The growth (OD600) of the WT and alr1 mutant in TSB containing (E) lactic acid (40 mM) (F) propionic acid (20 mM) and (G) itaconic acid (20 mM) in the presence or absence of 5 mM D-Ala.

The molecular docking results for lactate, propionate, and itaconate in the D-Ala binding site of Ddl also show similar types of interactions, but with variable poses and slight orientation differences compared to that observed for acetate in the crystal structure (Figure 6D). The D-Ala binding site, consisting of primarily charged and polar atoms, allows for a range of binding modes for these small anions, where the ligand size is a stronger factor in determining the binding location. Acetate and propionate, being smaller and less sterically hindered, bind preferentially near the Arg 291 side chain that coordinates the acid moiety of D-ala during the enzymatic reaction (Figure 6D). Meanwhile, itaconate and L-lactate bind in the more spacious region between Lys251 and Ser317 (Figure 6D). The Glide scores from the docking results, which provide a rough estimate of the ΔG of binding for each ligand suggest modest affinity to the identified binding sites (Table S4).

To determine if these organic acids could impact Ddl function, the WT and the alr1 mutant were challenged with lactic, propionic and itaconic acids (Figure 6E-G). All three organic acids inhibited the growth of the alr1 mutant. The addition of D-Ala to the culture media rescued the growth of the alr1 mutant to WT levels, (Figure 6E-G) consistent with Ddl being the target of lactate, propionate and itaconate. Moreover, overexpression of ddl in the alr1 mutant also restored growth of the alr1 mutant following the organic acid challenge (Figure 6-figure supplement 1A-C). These findings collectively suggest that various organic acid anions can inhibit Ddl activity in S. aureus.

Discussion

Intracellular anion accumulation has long been hypothesized to drive weak organic acid toxicity in bacteria (18, 29, 30). However, few studies have investigated the mechanism by which weak acid anions inhibit bacterial growth. Acetic acid is particularly interesting among weak acids, given that it is a common byproduct of glucose catabolism in bacteria and is excreted in high concentrations (31). S. aureus does not catabolize acetate as a carbon source unless glucose is first exhausted from its environment (32). Thus, in the presence of glucose, acetate can accumulate intracellularly in S. aureus as a function of the bacterial transmembrane pH gradient, especially when acetic acid concentrations are high in the immediate vicinity of cells. Here we determine that at high intracellular concentrations, acetate anions directly bind Ddl and inhibit D-Ala-D-Ala production to adversely impact peptidoglycan cross-linking (Figure 7).

Model depicting the role of Alr1 in countering organic acid anion-mediated inhibition of Ddl.

During its growth, S. aureus (WT) maintains a substantial intracellular pool of D-Ala through the activity of Alr1. Any excess D-Ala is subsequently converted into D-Glu by the action of the Dat enzyme. The high concentration of D-Ala is crucial for the optimal functioning of Ddl and serves to prevent the inhibition of Ddl by acetate (Ac-) and other organic acid anions. This process generates sufficient D-Ala-D-Ala, which is rapidly incorporated into the PG tripeptide precursor UDP-NAM-AEKAA to form UDP-NAM-AEKAA, which ultimately contributes to a robust cross-linked PG (murein) sacculus. In the alr1 mutant, the Dat reaction orientation is switched to preserve intracellular D-Ala. Nevertheless, this change is inadequate to maintain sufficient D-Ala pool to shield Ddl from inhibition by Ac-, due to tight control of dat translation. This results in an excess of UDP-NAM-AEK, which competes effectively with UDP-NAM-AEKAA for PG incorporation. The absence of a terminal D-Ala-D-Ala in the PG hinders crosslinking and leads to impaired growth following acetate intoxication.

However, S. aureus exhibits a remarkable tolerance to acetate intoxication due to the robust production of D-Ala by Alr1, which ultimately increases D-Ala-D-Ala pools (Figure 7).

Multiple lines of evidence demonstrate Ddl to be the target of acetate anions. First, LC- MS/MS analysis revealed that acetate intoxication decreased D-Ala-D-Ala pools, but not D-Ala in S. aureus, pointing to Ddl as the target of acetate. Second, DSF and in-vitro enzyme kinetic studies showed that acetate could bind and inhibit purified rDdl through a mixed inhibition mechanism. Third, structural analysis of the Ddl-inhibitor complex confirmed that acetate binds to both the ATP-binding and D-Ala binding sites within Ddl and further induced conformational changes to the dynamic ω loop, which weakens the binding of ATP to the Ddl active site. Finally, overexpression of ddl alone was sufficient to overcome acetate-mediated inhibition of the alr1 mutant and restore growth to WT levels.

Inhibitors that bind an enzyme’s catalytic substrate binding sites are usually competed out by high concentrations of substrates. However, acetate inhibits Ddl through a mixed inhibition mechanism despite binding to the substrate binding pockets of Ddl. We suspect this is due to additional conformational changes observed in the dynamic ω loop that affords more allosteric- like effects on enzyme activity. However, we cannot rule out that acetate might bind to additional sites in the Ddl-ATP complex, Ddl-ADP complex, or a Ddl-ADP-phospho-D-Ala complex with varying affinities. The differences in the temperature shifts observed in DSF with various substrate complexes support this possibility. The crystal structures of Ddl/acetate complexes with different substrates could provide a more precise conclusion about the inhibitory modality of Ddl by acetate. In line with acetate’s inhibitory effect on Ddl, we observed that acetate intoxication in the alr1 mutant led to a disproportionate increase in the cytosolic pool of PG tripeptide intermediate (UDP-NAM-AEK) compared to the pentapeptide form (UDP-NAM-AEKAA). Previous reports have suggested that MraY might facilitate the integration of UDP-NAM- tripeptide into S. aureus PG, especially when its concentration within cells exceeds that of UDP-NAM-pentapeptide (33, 34). Our findings strongly support this hypothesis, as the analysis of the alr1 mutant’s cell wall muropeptides revealed a clear elevation in the level of the disaccharide- tripeptide NAG-NAM-AEK. The inhibition of Ddl by acetate would further reduce the presence of terminal D-Ala-D-Ala moieties within alr1 muropeptides which likely leaves these cells incapable of withstanding the outward-directed cell turgor pressure, ultimately leading to cell death (34).

Despite acetate inhibiting Ddl through a mixed inhibition mechanism, it should be noted that a functional Alr1 or even the supplementation of D-Ala in culture media can provide significant tolerance against acetate intoxication in S. aureus. These observations suggest that Ddl is only weakly inhibited by acetate, which is also evident from the relatively high IC50 of approximately 400 mM observed in our kinetic experiments with S. aureus Ddl. The weak inhibition of Ddl would suggest that inflating the cytosolic D-Ala pools could promote sufficient generation of D-Ala-D-Ala to counter acetate intoxication. Indeed, it has been estimated that S. aureus maintains a high concentration of roughly 30 mM intracellular D-Ala (35), which we now demonstrate to be critical in countering acetate intoxication. High concentrations of acetate and other short-chain fatty acids are typically found in the human gut, where S. aureus can colonize (8–10, 12). In these environments, the robust production of D-Ala by staphylococcal Alr1 is likely just one mechanism by which S. aureus counters the inhibition of Ddl by weak organic acid anions. Additionally, S. aureus may also utilize D-Ala produced by the gut microbiota to minimize the impact of acetate intoxication on Ddl (36, 37).

The existence of pepV and dat within the same operon suggests that these genes may have evolved related functions. In Lactococcus lactis the PepV dipeptidase activity was shown to be important for supplying cells with L-Ala which was eventually incorporated into PG (38). In this context, pepV and dat may have a similar role in modulating the intracellular alanine pool. A surprising finding of our study was that dat expression is relatively stable and tightly controlled in S. aureus due to its SD motif being located within the coding region of pepV. Furthermore, such a genetic arrangement has been linked to translational coupling (39), wherein active translation from the first gene promotes the translation of the following gene in the operon, which in the case of dat was not sufficient to overcome acetate toxicity in the alr1 mutant. Two central mechanisms of translational coupling have been proposed. The first involves secondary and tertiary mRNA structures that either occlude or encompass the SD motif of downstream genes and shield it from ribosomes, thus preventing its translation (40). These mRNA structures can be relieved when a ribosome initiates translation from the first gene of the operon and exposes the downstream intragenic SD sequences to new 30S ribosomal subunits (34). In the second mechanism, continued translation of the first gene of the operon is necessary to increase the abundance of ribosomes in the TIR of the second gene, resulting in its enhanced translation (35). Irrespective of the mechanism of translational coupling, our results suggest that genetic arrangements that promote translational coupling might also limit the overall production of dat and thus prevent it from functionally complementing the alr1 mutant following acetate intoxication. Since our data suggest that Dat primarily promotes flux from D-Ala to D-Glu when Alr1 is active, the tight control of dat through translational coupling could prevent the depletion of the intracellular reserves of D-Ala necessary to overcome Ddl inhibition during acetate intoxication. Thus, the elevated D-Ala pool maintained within the cell could represent a strategic adaptation by S. aureus to combat Ddl inhibition caused by organic acids typically present in the niches colonized by this bacterium.

In conclusion, our findings demonstrate that Ddl is the primary target of acetate anion intoxication in S. aureus. However, other biologically relevant organic anions like lactate, propionate and itaconate could also inhibit the alr1 mutant similar to acetate. Furthermore, the growth inhibition of the alr1 mutant by these organic acids could be rescued following D-Ala supplementation, which suggests that Ddl is a bona fide and conserved target of various organic acid anions. Indeed, it is tempting to speculate that the robust Alr1 activity leading to the accumulation of millimolar levels of D-Ala may have evolved in part to offset the inhibition of Ddl from the toxic effects of organic anions.